---
title: "Looplook: An integrative suite for target assignment and functional annotation of chromatin interactions"
author: 
  - name: Ying ZHANG
    affiliation: Zhejiang University
    email: 12207129@zju.edu.cn
    comment: ORCID:0009-0005-9644-7062
  - name: Xingze HUANG
    affiliation: Zhejiang University
    email: 22407026@zju.edu.cn
    comment: ORCID:0009-0002-9286-1344
  - name: Liang XU
    affiliation: Zhejiang University
    email: xuliang.phd@zju.edu.cn
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{looplook: An Integrative Framework for Chromatin Interaction Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 7,
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
```

***
# Introduction

Welcome to **`looplook`**, a versatile R/Bioconductor toolkit developed to integrate three-dimensional (3D) chromatin architecture data (e.g., HiChIP, ChIA-PET, Hi-C) with other tabular omics datasets, including transcriptomics, chromatin accessibility, protein-DNA interactions (derived from ChIP-seq, CUT&Tag, or CUT&RUN), and genetic variants annotated by genome-wide association studies (GWAS).

Numerous studies have demonstrated that distal regulatory elements physically interact with target gene promoters via **3D chromatin looping**, thereby regulating the expression of genes located tens of kilobases to megabases away along the linear genome. However, conventional annotation strategies typically assign putative regulatory elements (e.g., informed by ChIP-seq peaks) to their **nearest neighboring genes in cis**, which often deviates from actual biological mechanisms. Consequently, the accurate assignment of non-coding genetic variants or orphan regulatory peaks to their **cognate target genes** represents a major bottleneck in the functional annotation of genomic regulatory elements.

To overcome this challenge, **`looplook`** systematically prioritizes experimentally validated spatial chromatin contacts to batch-annotate thousands of regulatory elements at a **genome-wide, high-throughput scale**, thereby identifying their candidate target genes with high confidence and unprecedented efficiency.

Beyond its utility as an integrative tool for target gene annotation, **`looplook`** serves as a **standalone platform** dedicated to chromatin loop analysis. Even in the absence of auxiliary omics data, this tool enables systematic annotation of the 3D chromatin interactome itself, classification of complex spatial interaction topologies (e.g., Enhancer-Promoter, Promoter-Promoter interactions), and quantification of node connectivity, which facilitates the identification of **dense regulatory hubs** and **enhancer cliques** (e.g., super-enhancers) that drive cell-type-specific transcriptional programs.

## Key Features & Capabilities

1. **The “3D Spatial Bridge” for Multi-Omics**: Integrates auxiliary genomic features (e.g., GWAS risk SNPs, ChIP-seq peaks) with 3D chromatin loops. It implements a rigorous “Smart Fallback” logic that prioritizes distal loop-mediated target genes while reverting to the nearest neighboring genes when no spatial interactions are detected, ensuring comprehensive and gapless target annotation.

2. **Comprehensive Loop Annotation & Topological Hub Detection**: In addition to loop mapping, **`looplook`** enables biologically informed annotation of the 3D chromatin interactome. It helps to categorize spatial interaction types and computes spatial node degrees to identify candidate 3D regulatory hubs.

3. **Expression-Aware Refinement**: Incorporates quantitative expressional data to filter out spurious structural contacts. As some chromatin regions may exhibit both enhancer and promoter activities, **`looplook`** provides a function to reclassify loop anchors which associate with transcriptionally silent reference promoters into **enhancer-like elements (eP)**, yielding a refined network of regulatory interactions.

4. **Reproducible Consolidation & Multi-source Consensus**: Utilizes graph-theoretic clustering to effectively harmonize biological replicates and mitigate potential technical noise. The framework can be used to identify multi-source consensus by integrating datasets from various experimental designs, such as HiChIP assays targeting different factors (e.g., **H3K27ac and Pol II**). This enables **orthogonal validation** to facilitate the construction of a high-confidence, unified 3D chromatin interactome based on consistent spatial features.

5. **Automated Multi-Omics Functional Interpretations**: Provides an end-to-end analytical engine for generating publication-grade visualizations and functional interpretations. This includes assessments of network connectivity and expression dynamics, transcription factor motif enrichment (**SeqLogos**), and pathway association networks (**Cnetplots**).

***

# Installation

To ensure full functionality, install `looplook` along with its recommended Bioconductor annotation dependencies:

```{r install, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Install annotation databases required for human (hg38) analysis
BiocManager::install(c("TxDb.Hsapiens.UCSC.hg38.knownGene", "org.Hs.eg.db"))

# Install looplook from GitHub
devtools::install_github("zying106/looplook")
```

```{r load_package}
library(looplook)
# Create a temporary directory to store all output plots and Excel files
out_dir <- tempdir()
```

***

# Module 1: Data Consolidation & Preprocessing

In 3D genomics analyses, individual replicates typically exhibit certain degrees of **inconsistency or noise**. The **`consolidate_chromatin_loops`** function serves as the foundational data-cleaning module, merging multiple replicates into a standardized, unified 3D chromatin interaction coordinate framework. 

This module identifies **multi-source consensus** by integrating datasets from varied experimental designs, such as HiChIP assays targeting H3K27ac and Pol II. By capturing shared spatial interactions across these inputs, the framework enables **orthogonal validation** to **mitigate potential technical artifacts** and assay-specific biases. This approach leverages consistent spatial features to provide a robust foundation for downstream analyses supported by multiple lines of experimental evidence.

### Parameter Strategy

* **`mode`**: Defines the overarching merging algorithm.
    * `"consensus"` *(Recommended)*: Implements graph-based connected component analysis to cluster spatially proximal anchors across samples. Only retains clusters detected in >= `min_consensus` biological replicates.
    * `"intersect"`: Enforces strict reference-based filtering, retaining loops that show full genomic overlap with the reference file (File 1). 
    * `"union"`: Retains all chromatin interactions across the entire cohort, ideal for exploratory pan-tissue analyses.
* **`min_consensus`**: When using `"consensus"` mode, this parameter defines the minimum number of biological replicates in which a loop cluster must be detected to be retained in the final dataset. If set to `NULL`, the algorithm dynamically computes a strict majority threshold (e.g., >= 75% of replicates).
* **`gap`**: Defines the maximum allowable spatial distance (in base pairs) between loop anchors for them to be considered as part of the same physical cluster (default: `1000`).
* **`min_raw_score` vs. `min_score` (The Dual-Filter)**:
    * `min_raw_score` acts as a pre-filter applied to individual BEDPE files before clustering (e.g., removing singleton noise where PET count < 2) to substantially reduce computational memory overhead.
    * `min_score` serves as a post-filter applied to the final merged chromatin interactome to ensure high-confidence interactions.
* **`blacklist_species`**: Automatically excludes chromatin loops overlapping with high-variance, artifact-prone genomic regions (e.g., centromeres, telomeres) by integrating the official ENCODE blacklist for specified species (e.g., `"hg38"`, `"mm10"`).
* **`region_of_interest`**: Accepts an auxiliary BED file (e.g., a specific disease-associated locus or ChIP-seq peak set) to exclude global background interactions, outputting only loops with physical connectivity to the target genomic region.

## Example 1: Building a Global Consensus Interactome

This represents the standard workflow for generating a high-confidence, whole-genome 3D chromatin interaction dataset via integration of biological replicates and removal of blacklist-associated artifacts.

```{r consolidate_global, message=FALSE, warning=FALSE}
# Locate example BEDPE replicates
f1 <- system.file("extdata", "example_loops_1.bedpe", package = "looplook")
f2 <- system.file("extdata", "example_loops_2.bedpe", package = "looplook")
global_out <- file.path(out_dir, "consensus_loops_global.bedpe")

# Execute consensus merging with strict Quality Control
consensus_global <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2, # Pre-filter sequencing noise
  blacklist_species = "hg38", # Apply artifact blacklist
  out_file = global_out
)
```

## Example 2: Feature-Driven Loop Filtering

In most scenarios, researchers are interested in 3D chromatin interactions associated with specific biological features, such as active enhancers (marked by H3K27ac) or particular transcription factors. 

By providing a BED file to `region_of_interest`, the tool enables efficient "on-the-fly" capture, retaining only those loops whose anchors overlap with the specified functional regions.

```{r consolidate_targeted, message=FALSE, warning=FALSE}
# Locate a BED file containing H3K27ac ChIP-seq peaks
h3k27ac_peaks <- system.file("extdata", "example_k27ac_peaks.bed", package = "looplook")
targeted_out <- file.path(out_dir, "H3K27ac_anchored_loops.bedpe")

# Execute consensus merging, strictly capturing H3K27ac-associated loops
consensus_targeted <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2,
  region_of_interest = h3k27ac_peaks, # Surgical capture via 1D features
  out_file = targeted_out
)
```

## Example 3: Multi-source Consensus for Orthogonal Validation

In integrative multi-omics analyses, orthogonal validation across independent functional assays is critical for identifying robust 3D regulatory interactions. For instance, enhancer-promoter (E-P) loops are typically supported by the spatial co-localization of H3K27ac (marking active chromatin) and RNA polymerase II (indicative of transcriptional machinery engagement). 
By utilizing the `"consensus"` mode to intersect independent loop datasets, `looplook` facilitates the identification of chromatin interactions supported by multi-source evidence. This rigorous intersection strategy effectively mitigates assay-specific technical biases and artifacts, yielding a refined set of topological hubs that couple physical chromatin architecture with active transcriptional regulation.


```{r consolidate_multi_modal, message=FALSE, warning=FALSE}
loop_k27ac <- system.file("extdata", "example_loops_H3K27ac.bedpe", package = "looplook")
loop_pol2 <- system.file("extdata", "example_loops_pol2.bedpe", package = "looplook")
dual_functional_out <- file.path(out_dir, "Dual_Functional_Consensus.bedpe")

consensus_dual <- consolidate_chromatin_loops(
  files = c(loop_k27ac, loop_pol2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2,
  out_file = dual_functional_out
)
```

***

### Module 2: 3D-Guided Annotation & Mapping

This module constitutes the core mapping engine of **`looplook`**. Using **`annotate_peaks_and_loops`**, users can classify the topological architecture of the interactome. Optionally, users may supply a `target_bed` (e.g., GWAS loci, eQTLs, ChIP-seq peaks) to trace non-coding regulatory signals to their putative functional target genes.

To resolve mapping ambiguities within densely populated gene loci, where a single loop anchor overlaps multiple candidate genes, the engine implements a rigorous three-step hierarchical annotation pipeline:

1. **Expression Pre-filter**: Eliminates transcriptionally silent genes based on the user-provided expression matrix.
2. **Functional Biotype Prioritization**: Prioritizes the remaining candidates by functional class in the following order: Protein Coding > lncRNA > Pseudogene.
3. **Dominant Expression Tiebreaker**: Designates the gene with the highest transcriptional abundance as the target gene to further resolve any remaining mapping ambiguities.

### Parameter Strategy & Core Inputs

This multi-omic integration relies on several key parameters:

* **`bedpe_file`**: The 3D structural framework (e.g., the high-confidence consensus interactome generated in Module 1).
* **`target_bed`**: The auxiliary genomic features of interest (e.g., a BED file of GWAS SNPs, ATAC-seq peaks, or ChIP-seq peaks) awaiting spatial target assignment.
* **`expr_matrix_file` & `sample_columns`**: Optional but strongly recommended. An expression matrix (e.g., RNA-seq) enables the Expression Pre-filter and Tiebreaker logic, substantially reducing false-positive gene assignments. 
* **`species`**: Specifies the genome assembly (e.g., `"hg38"`, `"mm10"`). The engine automatically loads the corresponding Bioconductor `TxDb` and `org.db` annotations.
* **`neighbor_hop`**: An advanced topological parameter for network traversal:
    * `0` *(Default)*: Restricts annotation to direct physical contacts (Anchor A ↔ Anchor B).
    * `1` *(Hub Mode)*: Evaluates secondary network effects within enhancer cliques (If Anchor A ↔ Anchor B ↔ Anchor C, this function enables topological linkage between A and C).
* **`tss_region`**: Defines the genomic window surrounding the transcription start site (TSS) used to define promoter regions (default: `c(-3000, 3000)` bp).

## Example A: Integrative Analysis (Loops + Genomic Features + RNA-seq)

This represents the comprehensive “3D Spatial Bridge” scenario. It links external genomic feature (peaks) to 3D‑inferred target genes, while simultaneously resolving ambiguous assignments in dense genomic regions by weighing gene expression.

```{r annotation_integrated, message=FALSE, warning=FALSE, results='hide'}
# Locate auxiliary genomic features and RNA-seq expression matrix
expr_path <- system.file("extdata", "example_tpm.txt", package = "looplook")
atac_path <- system.file("extdata", "example_peaks.bed", package = "looplook")
if (requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  res_integrated <- annotate_peaks_and_loops(
    bedpe_file = global_out, # Uses the global consensus network from Module 1
    target_bed = atac_path, # External genomic features
    expr_matrix_file = expr_path, # Activates the rigorous 3-step tiebreaker
    sample_columns = c("con1", "con2"),
    species = "hg38",
    neighbor_hop = 0, # Focus on direct physical contacts
    hub_percentile = 0.95, # Top 5% nodes defined as hubs
    out_dir = out_dir,
    project_name = "Example_HiChIP_Integrative"
  )
}
```

## Example B: Intrinsic Loop Profiling (Loops ONLY)

For users focused exclusively on 3D interactome architecture without auxiliary data related to genomic features, looplook  functions as a standalone platform for loop profiling and structural hub identification.


```{r annotation_loops_only, message=FALSE, warning=FALSE, results='hide'}
if (requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  res_loops_only <- annotate_peaks_and_loops(
    bedpe_file = global_out,
    target_bed = NULL, # Omit genomic features entirely
    expr_matrix_file = expr_path, # Resolves multi-gene conflicts at loop anchors
    sample_columns = c("con1", "con2"),
    species = "hg38",
    neighbor_hop = 0,
    out_dir = out_dir,
    project_name = "Example_Loops_Only"
  )
}
```

### Deep Dive: Output Data Dictionary

The function exports a comprehensive Excel workbook (`*_Basic_Results.xlsx`) containing multi-layered annotations. Interpretation of spatial regulatory logic depends critically on the hierarchical structure of its columns:

#### 1. `target_annotation` (The Genomic Feature-to-3D Mapping Result)
*(Generated only when `target_bed` is provided)*

* **`annotation`**: Genomic context of the input peaks (e.g., "Distal Intergenic", "Promoter").
* **`Linked_Loop_IDs`**: Identifiers of 3D loops that overlap the peak.
* **`All_Loop_Connected_Genes`**: All genes (including promoters and gene bodies) topologically linked to the peak via 3D looping.
* **`Regulated_promoter_genes`**: A stringent subset restricted to genes whose promoters are directly contacted by the peak via 3D looping. 
* **`Assigned_Target_Genes`**: Target genes assigned *exclusively* via 3D loops. If the peak does not overlap any structural loop, this value remains empty (`NA`).
* **`*_Filled` Columns (The "Smart Fallback" Logic)**: Columns suffixed with `_Filled` (e.g., `Assigned_Target_Genes_Filled`, `Regulated_promoter_genes_Filled`) provide complete, gapless annotations. For peaks within looped regions, they retain 3D-refined distal targets; for peaks in unlooped regions, they will be annotated to the nearest linear gene.

#### 2. `loop_annotation` (3D Network Architecture)

* **`loop_type`**: Topological classification (e.g., E-P, P-P, E-E) determined by the biotype-aware logic.
* **`All_Anchor_Genes`**: All genes located linearly within the left and right anchors of a given loop.
* **`Putative_Target_Genes`**: Biologically resolved target genes. For an E-P loop, this field will report only the gene at the Promoter anchor.
* **`Cluster_All_Genes`**: All genes within the broader 3D cluster or clique in which the loop resides, reflecting the extended topological neighborhood.

#### 3. `anchor_annotation` (Spatial Hub Footprints)

* **`Cluster_Locus_Genes`**: All genes from loop anchors within a spatially interconnected hub (defined by network-connected components), representing the full gene set associated with the 3D hub.

#### 4. `promoter_centric_stats` & `distal_element_stats` (Topological Hub Detection)

These metrics quantify interactome connectivity to identify candidate regulatory hubs from complementary perspectives:

**A. Gene-Centric (`promoter_centric_stats`)**

- **`Total_Loops` / `n_Linked_Promoters` / `n_Linked_Distal`**: Deconstructs a gene's 3D connectome. High `n_Linked_Distal` suggests potential regulation by enhancer clusters, whereas high `n_Linked_Promoters` may indicate possible participation in multi-gene transcription.
- **`Dominant_Interaction`**: The predominant topological class governing the gene (e.g., E-P vs. P-P).
- **`Is_High_Connectivity_Gene`** *(and Distal variant)*: Binary flags defined by the `hub_percentile` to prioritize putative 3D regulatory hubs.

**B. Enhancer-Centric (`distal_element_stats`)**

- **`Target_Genes`**: Promoters physically contacted by the distal anchor.
- **`n_Linked_Promoters`**: Facilitates identification of candidate master enhancers that contact multiple distinct genes.
- **`Is_High_Connectivity_Distal_Element`**: Flags highly connected non-coding elements that may act as 3D hubs.

### Deep Dive: Output Visualizations

The module automatically generates up to 10 publication-grade PDF plots, enabling multi-scale interrogation of the 3D interactome.

#### 1. Global 3D Network Profiling
*(Always Generated)*

* **Loop Type Proportions** (`*_Basic_Donut.pdf` & `*_Basic_Circular.pdf`): The Donut chart visualizes the raw physical frequency of each topological interaction. The Circular plot quantifies unique target genes per loop type, intuitively illustrating how certain loop types might disproportionately regulate target genes.

* **Karyotype Density Heatmaps** (`*_Basic_Karyo_Anchors.pdf` & `*_Basic_Karyo_LoopGenes.pdf`): Genome-wide ideograms illustrating the density of 3D anchors and their associated genes, highlighting architectural hotspots.

* **Topological Overlap** (`*_Basic_Flower.pdf`): A simplified flower plot showing the intersection of target genes (Core vs. Unique) across different loop types.
* **Anchor Genomic Distribution** (`*_Basic_Anchor_Genomic_Distribution.pdf`): Pie chart summarizing the genomic context of 3D anchors (e.g., Distal Intergenic, Intron, Promoter).

#### 2. Genomic Feature-to-3D Target Profiling
*(Requires `target_bed`)*

* **Target Connected Rose Chart** (`*_Target_Rose.pdf`): Coxcomb chart illustrating the dominant 3D topologies associated with user-defined genomic features.

* **Target Genomic Distribution Pies** (`*_Target_Genomic_Distribution.pdf` & `*_Target_Loop_Genomic_Distribution.pdf`): Compares the genomic distribution of input genomic features (peaks) versus those successfully linked via 3D loops. This comparison is insightful for demonstrating the enrichment of spatial interactions.
* **Target Genes Karyotype** (`*_Basic_Karyo_TargetGenes.pdf`): Chromosomal density map of final putative target genes, identifying functionally active multi-omics loci.

***

# Module 3: Expression-Aware Refinement

Physical proximity is a structural prerequisite, but not a direct proxy for active transcriptional regulation. This module integrates quantitative transcriptome data to systematically filter out transcriptionally silent physical chromatin contacts, ensuring only functionally relevant 3D interactions are retained.


### Parameter Strategy & Core Inputs

The `refine_loop_anchors_by_expression` function implements functional filtration via the following key parameters:

- **`annotation_res`**: Foundational 3D annotation output generated by Module 2.
- **`expr_matrix_file` & `sample_columns`**: Quantitative expression matrix (e.g., TPM, FPKM) and the specific replicates to average, establishing a robust baseline expression profile.
- **`threshold` & `unit_type`**: Defines the quantitative expression cutoff (e.g., `threshold = 1`, `unit_type = "TPM"`) required to consider a gene biologically active. This parameter enables downward compatibility with various normalization methods.
- **`reclassify_by_expression`**: A transformative logical parameter (`TRUE`/`FALSE`). When enabled, transcriptionally silent Promoters (`P`) and Gene Bodies (`G`) are not simply discarded; instead, they are reclassified as enhancer-like regulatory elements (`eP`, `eG`). This correction refines the regulatory topology (e.g., reclassifying a functionally silent `P-P` loop into a curated `eP-P` loop).

## Example A: Standard Filtration (Strict Removal)

In this scenario, a strict expression threshold eliminates loops where putative target genes are transcriptionally silent, without modifying the original structural classifications.

```{r refinement_basic, message=FALSE, warning=FALSE}
res_integrated <- system.file("extdata", "analysis_results.RData", package = "looplook")

if (res_integrated != "") {
  load(res_integrated)
  res_basic <- refine_loop_anchors_by_expression(
    annotation_res = res_integrated,
    expr_matrix_file = expr_path,
    sample_columns = c("con1", "con2"),
    threshold = 1.0,
    unit_type = "TPM",
    reclassify_by_expression = FALSE,
    out_dir = out_dir,
    project_name = "Example_Basic_Filter"
  )
}
```

## Example B: Expression-Aware Reclassification (Recommended)

This advanced mode is strongly recommended, as it preserves the valuable 3D structural framework while refining functional annotations. A silent promoter anchor is reclassified as an enhancer (`eP`), acknowledging that while its reference gene is transcriptionally inactive, this region itself may function as an active regulatory element for a distant gene.

```{r refinement_reclassified, message=FALSE, warning=FALSE, eval = FALSE}
if (exists("res_integrated")) {
  refined_res <- refine_loop_anchors_by_expression(
    annotation_res = res_integrated,
    expr_matrix_file = expr_path,
    sample_columns = c("con1", "con2"),
    threshold = 1.0,
    unit_type = "TPM",
    reclassify_by_expression = TRUE, # Activates biological reclassification (e.g., P -> eP)
    out_dir = out_dir,
    project_name = "Example_Reclassified_Filter"
  )
}
```

### Deep Dive: Filtration Visualizations

This module automatically generates a specialized suite of visual diagnostics to quantify transcriptome-guided refinement efficacy and characterize the surviving functional network:

#### 1. Global Filtration Profiling
*(Always Generated)*

- **Filtration Effect Dumbbell** (`*_Comparison_Dumbbell.pdf`): Quantifies the "cleaning power" of the expression threshold by visualizing the numerical contrast between original structural loops and surviving functional loops across all topological classes, highlighting the importance of removing untranscribed structural noise.

- **Refined Loop Proportion Rose** (`*_Rose.pdf`): Coxcomb chart illustrating the topological distribution (by count) of the interactome after expression-guided reclassification and filtration.
- **Refined Active Genes Karyotype** (`*_Refined_Karyo_Active.pdf`): Genome-wide ideograms mapping the chromosomal density of transcriptionally active loop-associated genes that passed the expression threshold. 

#### 2. Genomic Feature-to-3D Target Filtration Profiling
*(Generated only if `target_bed` was provided in Module 2)*

- **Multi-Omics Sankey Diagram** (`*_Target_Sankey.pdf` / `.html`): Core integration diagnostic that maps the fate of inputted genomic features across three logical tiers: Genomic Region (L1) → 3D Loop Connection (L2) → Target Expression Status (L3). This intuitively reveals the proportion of GWAS/ChIP-seq peaks structurally connected to functionally active genes.

- **Refined Target Loop Donut** (`*_Target_Loop_Donut.pdf`): Visualizes the topological distribution of functionally active loops that bridge user-defined genomic features.
- **Refined Target Genes Karyotype** (`*_Refined_Karyo_TargetGenes.pdf`): Chromosomal density map of strictly refined putative target genes, facilitating visualization of transcriptionally verified multi-omics loci.

***

## Module 4: Automated Functional Profiling

Following identification of high-confidence putative target genes, `profile_target_genes` provides a fully automated, end-to-end multi-omics analysis pipeline. It integrates 3D genomic interactions with transcriptomic data to systematically characterize the functional landscape and regulatory mechanisms underlying the identified target genes.



### Parameter Strategy: A Highly Modular Pipeline

This function supports highly flexible experimental design through a set of routing parameters that define the gene set to be analyzed, as well as toggle parameters that enable or disable specific downstream biological modules.

#### 1. Core Data Inputs

- **`annotation_res`**: The foundational annotation result, inherited directly from Module 2 or the refined output from Module 3.
- **`diff_file` & `lfc_col`**: Differential expression summary (e.g., from DESeq2) required for fold change profiling and gene set enrichment analysis (GSEA).
- **`expr_matrix_file` & `metadata_file`**: Normalized expression matrix and sample metadata, which are essential for generating heatmaps and Raincloud plots.

#### 2. Target Selection & Mapping Strategies

This section contains the most critical routing parameters that define the biological scope and stringency of downstream analyses:

- **`target_source`** (The Biological Scope):
    - `"targets"` *(Variant/Peak-Centric)*: Focuses exclusively on the putative genes regulated by user-defined genomic features (e.g., GWAS SNPs, ATAC-seq peaks).
    - `"loops"` *(Global Network-Centric)*: Analyzes the full 3D interactome, including all genes associated with spatial loops, without considering user-defined genomic features.
    - `c("loops", "targets")`: Analyzes both scopes simultaneously, which is strongly recommended for comparative functional profiling.
- **`target_mapping_mode`**:
    - `"all"` *(Default)*: Retains broad 3D regulatory targets, including distal enhancers looping to gene bodies.
    - `"promoter"`: Applies stringent filtering, requiring loops to anchor explicitly at canonical promoter regions.
- **`include_Filled`** (The Stringency Toggle): Logical parameter controlling the purity of spatial regulation:
    - `TRUE` *(Hybrid Mode)*: Utilizes the comprehensively merged annotation (`_Filled` columns from Module 2). It prioritizes 3D loop-derived targets while rescuing unlooped genomic features (peaks) by assigning them the nearest genes, ensuring no data loss and a complete functional overview.
    - `FALSE` *(Pure Spatial Mode)*: Highly stringent. It only keeps 3D loop-derived targets with clear annotation.

- **`use_nearest_gene`**: Logical parameter. If `TRUE`, the pipeline bypasses 3D loop-based assignment and uses only the nearest neighboring gene, serving as a conventional baseline reference for Control comparisons.

#### 3. Downstream Functional Analyses

- **`run_go`**: Executes Gene Ontology (Biological Process, BP) enrichment and generates Divergent Concept Networks.
- **`run_motif` & `genome_id`**: Scans proximal and distal anchor sequences against JASPAR core motifs (requires `BSgenome`).
- **`run_ppi` & `ppi_score`**: Constructs Protein-Protein Interaction (PPI) networks via the STRING database.

## Example A: Comprehensive Integrative Profiling (Recommended)

Ideal for complete multi-omics integration (e.g., HiChIP + ATAC-seq). This mode comprehensively maps and profiles target genes of interest.

```{r profiling_scenario_A, message=FALSE, warning=FALSE}
diff_path <- system.file("extdata", "example_deg.txt", package = "looplook")
meta_path <- system.file("extdata", "example_coldata.txt", package = "looplook")

if (exists("refined_res")) {
  res_A <- profile_target_genes(
    annotation_res = res_integrated,
    diff_file = diff_path,
    lfc_col = "log2FoldChange",
    expr_matrix_file = expr_path,
    metadata_file = meta_path,
    target_source = c("loops", "targets"), # Analyzes both sources
    target_mapping_mode = "all",
    include_Filled = TRUE,
    use_nearest_gene = FALSE,
    project_name = "Scenario_A_Integrative",
    out_dir = out_dir,
    run_motif = FALSE, # Set TRUE in real analysis to scan JASPAR motifs
    run_go = FALSE, # Set TRUE in real analysis for pathway enrichment
    run_ppi = FALSE
  )
}
```

## Example B: Peak-Driven Strict Promoter Profiling (High Stringency)

This mode applies a rigorous dual-filtering strategy to yield a high-confidence target gene set.

First, `target_source = "targets"` restricts profiling to genes regulated specifically by user-defined genomic features (e.g., GWAS SNPs or TF binding sites), excluding background spatial loops. Second, `target_mapping_mode = "promoter"` enforces topological stringency: loops linking peaks to target genes must directly contact canonical or reference promoters (i.e., strict `E-P` or `P-P` loops), excluding more permissive connections such as enhancer-gene body (`E-G`) interactions.

```{r profiling_scenario_B, message=FALSE, warning=FALSE, eval = FALSE}
if (exists("refined_res")) {
  res_B <- profile_target_genes(
    annotation_res = res_integrated,
    diff_file = diff_path,
    lfc_col = "log2FoldChange",
    expr_matrix_file = expr_path,
    metadata_file = meta_path,
    target_source = "targets", # 1. Focus exclusively on inputted genomic features
    target_mapping_mode = "promoter", # 2. Require strict loop-to-promoter collision
    include_Filled = TRUE,
    use_nearest_gene = FALSE,
    project_name = "Scenario_B_StrictPromoter",
    out_dir = out_dir,
    run_motif = FALSE, # Set TRUE in real analysis to scan JASPAR motifs
    run_go = FALSE, # Set TRUE in real analysis for pathway enrichment
    run_ppi = FALSE
  )
}
```

## Example C: 1D Linear Annotation Baseline (The Control)

By setting `use_nearest_gene = TRUE`, this mode intentionally bypasses the 3D spatial topology and assigns features exclusively to their nearest neighboring gene. It serves as a conventional “baseline control.” Comparing results against Example A or B enables rigorous demonstration of the novel functional insights gained from 3D chromatin looping analyses relative to conventional method.

```{r profiling_scenario_C, message=FALSE, warning=FALSE, eval = FALSE}
if (exists("refined_res")) {
  res_C <- profile_target_genes(
    annotation_res = res_integrated,
    diff_file = diff_path,
    lfc_col = "log2FoldChange",
    expr_matrix_file = expr_path,
    metadata_file = meta_path,
    target_source = "targets", # Focus on inputted peaks
    target_mapping_mode = "all",
    include_Filled = FALSE,
    use_nearest_gene = TRUE, # Nearest neighboring gene (The Control)
    project_name = "Scenario_C_LinearControl",
    out_dir = out_dir,
    run_motif = FALSE, # Set TRUE in real analysis to scan JASPAR motifs
    run_go = FALSE, # Set TRUE in real analysis for pathway enrichment
    run_ppi = FALSE
  )
}
```

### Deep Dive: Functional Visualizations

This module executes an extensive downstream pipeline that generates a complete, publication-ready figure panel. Rather than simple descriptive summaries, these visualizations are designed to test research hypotheses related to multi-omics integration.

#### 1. Expression & Topological Dynamics

- **Transcriptional Profiling** (`*_LFC_Violin.pdf` & `*_Expression_Heatmap.pdf`): Integrates quantitative transcriptome data (e.g., RNA-seq) and differential expression statistics to evaluate whether 3D-annotated target genes exhibit statistically significant expression shifts relative to the global genome background. This provides a quantitative measure of functional relevance and correlative evidence of regulatory modulation.
- **Topological-Transcriptional Association** (`*_Connectivity_Scatter.pdf` & `*_Raincloud_*.pdf`): Cross-references 3D connectivity (e.g., number of linked enhancers) with transcript abundance and fold changes. This may help to test the "hub hypothesis" (whether increased spatial connectivity correlates with more dynamic transcriptional output).

#### 2. Regulatory Motif Scanning

- **Spatially Asymmetric Motif Signatures** (`*_Motif_*_RankScatter.pdf` & `*_Logos.pdf`): Separately analyses transcription factor motifs enriched in the promoter-proximal anchors and distal (enhancer) anchors.

#### 3. Pathway & Network Enrichment

- **Spatial Gene Set Enrichment** (`*_GSEA.pdf`): Projects 3D-derived target gene sets onto the globally ranked differential expression landscape. Rather than relying on pre-defined pathways, it treats the specific spatial network as a biologically meaningful gene set, and examines whether the 3D interactome contributes to the observed transcriptomic phenotype.
- **Core Effector Networks** (`*_GO_Network.pdf`, `*_GO_Lollipop_*.pdf` & `*_PPI_Network_*.pdf`): Derived from Gene Ontology and STRING analyses. Rather than simply listing enriched terms, the Divergent Concept Networks and PPI graphs link key pathways to high-magnitude differentially expressed hub genes, bridging spatial chromatin architecture, transcriptional regulation, and functional protein networks.

***

## Module 5: IGV-Style Track Visualization

To visualize the local spatial interactome with high resolution, this module generates an IGV-style multi-tiered genomic browser view via the `plot_peaks_interactions` function. This visualization integrates multi-omics datasets into three distinct, non-overlapping genomic tracks, enabling intuitive assessment of 3D structural context:



- **Loop Track (Top)**: Depicts 3D chromatin interactions as Bezier arcs connecting paired spatial anchors. When `score_to_alpha = TRUE` is enabled, the opacity (alpha channel) of each arc is scaled proportionally to the quantitative interaction score.
- **Target Track (Middle)**: Renders user-provided genomic features (e.g., ChIP-seq peaks or GWAS SNPs) from the input BED file, facilitating direct visualization of their spatial overlap with the loop anchors.
- **Gene Track (Bottom)**: Retrieves the longest canonical transcript for each gene using standard Bioconductor annotation databases, with exons and introns mapped according to strand directionality. Overlapping gene models are vertically separated to avoid label collisions and improve readability.

### Parameter Strategy & Core Inputs

The function provides fine-grained control over genomic locus selection, data filtering, and visual aesthetics through the following arguments:

#### 1. Data Inputs & Genomic Coordinates

- **`bedpe_file`**: Character. Specifies the path to the BEDPE-format spatial interaction file. The function utilizes the first six columns for paired anchor coordinates and the seventh column (if present) as the quantitative interaction score.
- **`target_bed`**: Character (Optional). Specifies the path to a user-supplied genomic feature file in standard BED format, rendered in the middle track to assess overlap with 3D anchors.
- **`chr`, `from`, `to`**: Specifies the exact genomic window for visualization. If `chr` is omitted (`NULL`), the function defaults to the chromosome with the highest loop count in the BEDPE file.
- **`min_score`**: Numeric (Optional). Applies a threshold to filter the displayed loops based on the interaction score column.

#### 2. Annotation Dependencies

- **`species`**: Character. Specifies the reference genome assembly (supported: `"hg38"`, `"hg19"`, `"mm10"`, or `"mm9"`). This parameter triggers automatic loading of the corresponding Bioconductor `TxDb` and `OrgDb` annotation packages for gene track rendering.

#### 3. Aesthetic Mapping & Track Configuration

- **`score_to_alpha`**: Logical flag. When `TRUE`, the quantitative interaction score is mapped to the transparency of Bezier arcs, enabling visual differentiation of interaction strength.
- **`max_levels`**: Integer. Defines the maximum vertical stacking limit for the loop arcs to manage rendering height and readability.
- **`base_anchor_height`**: Numeric. Sets the vertical thickness of the rectangular anchors plotted at the base of the loop arcs.
- **Color Controls** (`loop_color`, `anchor_color`, `overlap_color`, `exon_color`, `intron_color`): Character strings specifying custom colors for structural elements across all three tracks.
- **`save_file`**: Character (Optional). Specifies the output file path and format (e.g., `".pdf"`) for exporting the final ggplot object.

```{r visual_track}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  # If 'from' and 'to' are omitted, it automatically detects the densest viewport.
  track_plot <- plot_peaks_interactions(
    bedpe_file = f1,
    target_bed = atac_path,
    chr = "chr1",
    from = 11884299,
    to = 12106581,
    species = "hg38",
    save_file = file.path(out_dir, "Locus_Track.pdf")
  )
}
```

***

# Session Info

```{r session_info, eval = TRUE}
sessionInfo()
```
