---
title: "Looplook: A multi-omics suite for expression-aware target assignment and topological profiling of 3D chromatin networks"
author: 
  - name: Ying Zhang
    affiliation: Zhejiang University
    email: 12207129@zju.edu.cn
    comment: ORCID:0009-0005-9644-7062
  - name: Xingze Huang
    affiliation: Zhejiang University
    email: 22407026@zju.edu.cn
    comment: ORCID:0009-0002-9286-1344
  - name: Xu Liang
    affiliation: Zhejiang University
    email: xuliang.phd@zju.edu.cn
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{looplook: An Integrative Framework for Chromatin Interaction Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 7,
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
```

***
# Introduction

Welcome to **`looplook`**, a highly sophisticated R/Bioconductor toolkit designed to decipher the complex interplay between **3D chromatin architecture** (e.g., HiChIP, ChIA-PET, Hi-C) and **1D multi-omics profiles** (including transcriptomics, chromatin accessibility, protein-DNA interactions via ChIP-seq/CUT&Tag, and genetic variants like GWAS SNPs).

The toolkit resolves a major bottleneck in functional genomics: accurately assigning non-coding variants or orphan peaks to their target genes. While traditional annotations rely on the *nearest linear gene* heuristic, this often fails to capture reality. Many distal regulatory elements physically contact target gene promoters via 3D chromatin interactions (chromatin looping), thereby regulating the expression of genes located tens of kilobases (kb) to megabases (Mb) away.

Furthermore, while genome browsers like IGV allow researchers to visually inspect these 3D connections at individual loci, such manual, low-throughput approaches are unfeasible for global analysis. `looplook` computationally bridges this critical 1D-to-3D gap at a **genome-wide, high-throughput scale**. It systematically prioritizes physical spatial contacts to batch-annotate thousands of regulatory elements, revealing their **high-confidence putative target genes** with unprecedented efficiency.

Beyond its power as a spatial bridge, `looplook` serves as a standalone powerhouse for **intrinsic loop profiling**. Even without auxiliary 1D inputs, it systematically annotates the 3D interactome itself, classifying complex spatial topologies (e.g., Enhancer-Promoter, Promoter-Promoter) and calculating node connectivities to uncover dense **regulatory hubs** and super-enhancer cliques that drive cell-type-specific transcriptional programs.

### Key Features & Capabilities

1. **The "3D Spatial Bridge" for Multi-Omics:** Unites auxiliary 1D genomic features (GWAS risk SNPs, ChIP-seq binding sites) with 3D loops. It features a rigorous "Smart Fallback" logic—prioritizing loop-assigned distal targets while gracefully falling back to the nearest linear genes if no spatial structure is found, ensuring comprehensive, gapless coverage.
2. **Comprehensive Loop Annotation & Topological Hub Detection:** Goes beyond treating loops as mere bridges by biologically annotating the 3D interactome itself. It precisely classifies spatial interactions and computes spatial node degrees to identify candidate 3D regulatory hubs.
3. **Expression-Aware Refinement:** Integrates quantitative RNA-seq data to filter false-positive structural interactions. It introduces the biological reclassification of mathematically silent promoters into enhancer-like elements (e.g., correcting a silent P-P loop to a functional eP-P loop), yielding a strictly active regulatory network.
4. **Reproducible Consolidation:** Employs graph-theoretic clustering to merge biological replicates and eliminate technical noise, forging a high-confidence consensus 3D interactome.
5. **Automated Multi-Omics Profiling:** A push-button engine that translates structural coordinate sets into publication-quality visualizations and functional insights, spanning network connectivity vs. expression dynamics, motif enrichment (SeqLogos), and pathway concept networks (Cnetplots).

***

# Installation

To ensure full functionality, install `looplook` along with its recommended Bioconductor annotation dependencies:

```{r install, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Install annotation databases required for human (hg38) analysis
BiocManager::install(c("TxDb.Hsapiens.UCSC.hg38.knownGene", "org.Hs.eg.db"))

# Install looplook from GitHub
devtools::install_github("zying106/looplook")
```

```{r load_package}
library(looplook)
# Create a temporary directory to store all output plots and Excel files
out_dir <- tempdir()
```

***

# Module 1: Data Consolidation & Preprocessing

In 3D genomics, individual replicates often suffer from high noise and low signal. The `consolidate_chromatin_loops` function acts as the foundational data-cleaning engine, merging multiple replicates into a clean, unified 3D coordinate framework.

## Parameter Strategy

* `mode`: Defines the overarching merging algorithm.
  * `"consensus"` *(Recommended)*: Uses graph-based connected component analysis to group nearby anchors across samples. Retains clusters present in $\ge$ `min_consensus` replicates.
  * `"intersect"`: Strict reference-based filtering (requires absolute overlap with File 1).
  * `"union"`: Retains all interactions across the cohort (useful for exploratory pan-tissue analysis).
* `min_consensus`: When using "consensus" mode, this defines the exact number of biological replicates a loop cluster must appear in to be retained. If left as `NULL`, the engine dynamically calculates a strict majority threshold (e.g., $\ge$ 75% of replicates).
* `gap`: The maximum spatial distance (in base pairs) allowed between anchors to consider them part of the same physical cluster (default: `1000`).
* `min_raw_score` vs. `min_score` (The Dual-Filter): 
  * `min_raw_score` is a pre-filter applied to individual BEDPE files *before* clustering (e.g., removing singleton noise where PET count < 2) to drastically reduce memory overhead. 
  * `min_score` acts as a post-filter applied to the *final merged interactome*.
* `blacklist_species`: Automatically drops loops falling into high-variance, artifact-prone genomic regions (e.g., centromeres, telomeres). By specifying `"hg38"`, `"hg19"`, `"mm10"`, or `"mm9"`, the engine applies the official ENCODE blacklist seamlessly.
* `region_of_interest`: A surgical precision tool. By providing an auxiliary BED file (e.g., a specific 2Mb disease locus like the *MYC* neighborhood), the engine drops the global background and strictly outputs loops physically connected to this targeted zone.

## Example 1: Building a Global Consensus Interactome

This is the standard workflow to generate a clean, whole-genome 3D backbone by merging replicates and eliminating blacklisted artifacts.

```{r consolidate_global, message=FALSE, warning=FALSE}
# Locate example BEDPE replicates
f1 <- system.file("extdata", "example_loops_1.bedpe", package = "looplook")
f2 <- system.file("extdata", "example_loops_2.bedpe", package = "looplook")
global_out <- file.path(out_dir, "consensus_loops_global.bedpe")

# Execute consensus merging with strict Quality Control
consensus_global <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2, # Pre-filter sequencing noise
  blacklist_species = "hg38", # Apply ENCODE artifact blacklist
  out_file = global_out
)
```

## Example 2: Feature-Driven Loop Filtering

Often, researchers are only interested in 3D interactions mediated by specific biological features, such as active enhancers (marked by H3K27ac) or specific transcription factors. By providing a ChIP-seq BED file to `region_of_interest`, the engine performs an elegant "on-the-fly" capture, discarding any loops that do not physically anchor at these functional sites.

```{r consolidate_targeted, message=FALSE, warning=FALSE}
# Locate a BED file containing H3K27ac ChIP-seq peaks
h3k27ac_peaks <- system.file("extdata", "example_k27ac_peaks.bed", package = "looplook")
targeted_out <- file.path(out_dir, "H3K27ac_anchored_loops.bedpe")

# Execute consensus merging, strictly capturing H3K27ac-associated loops
consensus_targeted <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2,
  region_of_interest = h3k27ac_peaks, # Surgical capture via 1D features
  out_file = targeted_out
)
```

## Example 3: Cross-Modality Consensus for Orthogonal Validation

In integrative multi-omics studies, orthogonal validation across distinct functional assays is critical for defining robust 3D regulatory interactions. For instance, bona fide Enhancer-Promoter (E-P) loops are characterized by the concurrent spatial proximity of H3K27ac (active enhancer/promoter regions) and RNA Polymerase II (transcriptional machinery). By utilizing the `"consensus"` mode to intersect these distinct loop datasets, `looplook` facilitates the extraction of multi-modally supported interactions. This stringent merging strategy may effectively mitigate assay-specific technical artifacts, yielding a refined set of topological hubs that strictly couple physical chromatin architecture with active transcriptional regulation.

```{r consolidate_multi_modal, message=FALSE, warning=FALSE}
loop_k27ac <- system.file("extdata", "example_loops_H3K27ac.bedpe", package = "looplook")
loop_pol2  <- system.file("extdata", "example_loops_pol2.bedpe", package = "looplook")
dual_functional_out <- file.path(out_dir, "Dual_Functional_Consensus.bedpe")

  consensus_dual <- consolidate_chromatin_loops(
    files = c(loop_k27ac, loop_pol2),
    mode = "consensus", 
    gap = 1000,
    min_raw_score = 2,
    out_file = dual_functional_out
  )
```

***

# Module 2: 3D-Guided Peak Annotation & Mapping

This module forms the core mapping engine of `looplook`. Using `annotate_peaks_and_loops`, users can classify the 3D topology of the interactome and, optionally, inject a `target_bed` (e.g., GWAS loci, eQTLs, ChIP-seq peaks) to trace non-coding signals to their **putative functional target genes**. 

Concurrently, to resolve mapping conflicts in dense gene loci (e.g., when a single loop anchor overlaps multiple genes), the engine seamlessly executes a rigorous 3-step hierarchical pipeline:

1. **Expression Pre-filter:** Instantly eliminates transcriptionally silent candidates based on your RNA-seq matrix.
2. **Functional Biotype Prioritization:** For the remaining active genes, it biologically prioritizes meaningful functional classes (Protein Coding > lncRNA > Pseudogene).
3. **Dominant Expression Tiebreaker:** If a mapping tie still exists (e.g., multiple active protein-coding genes within the same anchor), it automatically resolves the conflict by selecting the dominant gene with the highest transcriptional abundance.

## Parameter Strategy & Core Inputs

To execute this intricate multi-omics integration, the function relies on several critical parameters:

* `bedpe_file`: The 3D structural backbone (e.g., the high-confidence consensus interactome generated in Module 1).
* `target_bed`: The 1D auxiliary features of interest (e.g., a BED file of GWAS SNPs, ATAC-seq peaks, or ChIP-seq binding sites) that require spatial target assignment.
* `expr_matrix_file` & `sample_columns`: While optional, providing an RNA-seq matrix allows the engine to activate the **Expression Pre-filter** and **Tiebreaker** logic, drastically reducing false-positive gene assignments.
* `species`: Specifies the genome assembly (e.g., `"hg38"`, `"mm10"`). The engine automatically detects and loads the appropriate Bioconductor `TxDb` and `org.db` annotations.
* `neighbor_hop`: A highly advanced topological parameter for network traversal:
  * `0` *(Default)*: Strictly annotates direct physical contacts (Anchor A touches Anchor B).
  * `1` *(Hub Mode)*: Evaluates secondary network effects within super-enhancer cliques (Anchor A touches B, B touches C $\rightarrow$ A is topologically linked to C).
* `tss_region`: Defines the spatial boundary of a promoter relative to the Transcription Start Site (TSS) (default: `c(-3000, 3000)` bp).

## Example A: Integrative Analysis (Loops + 1D Peaks + RNA-seq)

This is the comprehensive "3D Spatial Bridge" scenario. It maps 1D external peaks to 3D target genes while simultaneously resolving gene locus conflicts using background expression data.

```{r annotation_integrated, message=FALSE, warning=FALSE}
# Locate auxiliary 1D peaks and RNA-seq expression matrix
expr_path <- system.file("extdata", "example_tpm.txt", package = "looplook")
atac_path <- system.file("extdata", "example_peaks.bed", package = "looplook")
if (requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  res_integrated <- annotate_peaks_and_loops(
    bedpe_file = global_out, # Uses the global consensus network from Module 1
    target_bed = atac_path, # 1D External features to map
    expr_matrix_file = expr_path, # Activates the rigorous 3-step tiebreaker
    sample_columns = c("con1", "con2"),
    species = "hg38",
    neighbor_hop = 0, # Focus on direct physical contacts
    hub_percentile = 0.95, # Top 5% nodes defined as hubs
    out_dir = out_dir,
    project_name = "Example_HiChIP_Integrative"
  )
}
```

## Example B: Intrinsic Loop Profiling (Loops ONLY)

If you are purely interested in the 3D interactome and do not have auxiliary 1D peaks, `looplook` serves as a standalone powerhouse for intrinsic loop profiling and structural hub detection. 


```{r annotation_loops_only, message=FALSE, warning=FALSE}

if (requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  res_loops_only <- annotate_peaks_and_loops(
    bedpe_file = global_out, 
    target_bed = NULL, # Omit 1D features entirely
    expr_matrix_file = expr_path, # Resolves multi-gene conflicts at loop anchors
    sample_columns = c("con1", "con2"),
    species = "hg38",
    neighbor_hop = 0,
    out_dir = out_dir,
    project_name = "Example_Loops_Only"
  )
}
```

## Deep Dive: Output Data Dictionary

The function seamlessly exports an extensively detailed Excel workbook (`*_Basic_Results.xlsx`). Understanding the hierarchical logic of these columns is critical to interpreting the spatial regulation:

### 1. `target_annotation` (The 1D-to-3D Mapping Result)
*(Generated only if `target_bed` is provided)*

* **`annotation`**: The local 1D genomic context of the peak (e.g., "Distal Intergenic", "Promoter").
* **`Linked_Loop_IDs`**: IDs of the specific 3D loops physically overlapping this peak.
* **`All_Loop_Connected_Genes`**: The broadest spatial reach. All genes (including both Promoters and Gene Bodies) topologically connected to this peak via the 3D network.
* **`Regulated_promoter_genes`**: A biologically stricter subset. It restricts the spatial connections *only* to genes where the peak loops directly to their active Promoters.
* **`Assigned_Target_Genes`**: Target genes derived *exclusively* via 3D physical loops. If the peak does not overlap any structural loop, this value remains empty (NA).
* **`*_Filled` Columns (The "Smart Fallback" Logic)**: Columns ending in `_Filled` (e.g., **`Assigned_Target_Genes_Filled`**, **`Regulated_promoter_genes_Filled`**) provide a **comprehensive, gapless annotation**. They retain the loop-derived distal targets if a spatial loop exists, but intelligently "fill" the empty rows by falling back to the nearest linear gene (`SYMBOL`) for peaks residing in unlooped regions.

### 2. `loop_annotation` (The 3D Network Architecture)

* **`loop_type`**: Topological classification (e.g., E-P, P-P, E-E) determined by the biotype-aware logic.
* **`All_Anchor_Genes`**: The strict physical footprint. All genes located linearly within the left and right anchors of this specific individual loop.
* **`Putative_Target_Genes`**: The biologically resolved target (e.g., For an E-P loop, it strictly outputs the gene located at the Promoter anchor, deliberately ignoring genes at the Enhancer anchor).
* **`Cluster_All_Genes`**: The macro-neighborhood. Because multiple loops often form dense complex cliques (Clusters), this column reveals the entire pool of genes involved in the broader 3D cluster that this specific loop belongs to.

### 3. `anchor_annotation` (The Spatial Hub Footprints)

* **`Cluster_Locus_Genes`**: All genes residing linearly within the merged 1D footprint of the spatial anchor cluster, defining the foundational baseline of the locus before 3D tracking.

### 4. `promoter_centric_stats` & `distal_element_stats` (Topological Hub Detection)

Quantifies the structural node degrees of the interactome to identify putative regulatory hubs from two complementary perspectives:

#### A. Gene-Centric (`promoter_centric_stats`)

* **`Total_Loops` / `n_Linked_Promoters` / `n_Linked_Distal`**: Deconstructs a gene's 3D connectome. A high `n_Linked_Distal` might suggest regulation by a super-enhancer cluster, while a high `n_Linked_Promoters` could indicate possible participation in a multi-gene transcription factory.
* **`Dominant_Interaction`**: The predominant topology orchestrating the gene (e.g., E-P vs. P-P).
* **`Is_High_Connectivity_Gene`** *(and Distal variant)*: Binary flags governed by the `hub_percentile` to help prioritize putative 3D regulatory hubs.

#### B. Enhancer-Centric (`distal_element_stats`)

* **`Target_Genes`**: Promoters physically contacted by the specific distal anchor.
* **`n_Linked_Promoters`**: Aids in exploring candidate "master enhancers" that might simultaneously interact with multiple distinct genes.
* **`Is_High_Connectivity_Distal_Element`**: Flags dense non-coding nodes, highlighting putative enhancer domains that may maintain the local 3D architecture.

## Deep Dive: Output Visualizations

The module auto-generates a comprehensive suite of up to 10 publication-grade PDF plots to evaluate the interactome from macro-chromosomal scales down to specific topological classes.

### 1. Global 3D Network Profiling
*(Always Generated)*

* **Loop Type Proportions (`*_Basic_Donut.pdf` & `*_Basic_Circular.pdf`)**: The Donut chart visualizes the raw physical frequency of topological interactions, demonstrating the dominance of P-P and G-P loops. The Circular plot goes further by quantifying the unique target genes orchestrated by each type, intuitively illustrating how certain loop types might disproportionately regulate a massive number of genes.
* **Karyotype Density Heatmaps (`*_Basic_Karyo_Anchors.pdf` & `*_Basic_Karyo_LoopGenes.pdf`)**: Macro-scale chromosomal ideograms that plot the physical density of 3D anchors and their natively associated loop genes across the genome, highlighting broad structural hotspots.
* **Topological Overlap (`*_Basic_Flower.pdf`)**: A simplified flower plot revealing the intersection of target genes (Core vs. Unique) across different loop types.
* **Anchor Genomic Distribution (`*_Basic_Anchor_Genomic_Distribution.pdf`)**: A pie chart mapping the raw 3D anchors to their local 1D genomic context (e.g., Distal Intergenic, Intron, Promoter), verifying the overall structural distribution of the spatial data.

### 2. 1D-to-3D Target Profiling
*(Requires `target_bed`)*

* **Target Connected Rose Chart (`*_Target_Rose.pdf`)**: A coxcomb chart that helps reveal which specific 3D topologies are predominantly utilized by your inputted 1D variants.
* **Target Genomic Distribution Pies (`*_Target_Genomic_Distribution.pdf` & `*_Target_Loop_Genomic_Distribution.pdf`)**: Compares the baseline 1D distribution of all input targets against the subset of targets successfully bridged by 3D loops. This direct comparison is crucial for demonstrating the enrichment of spatial interactions.
* **Target Genes Karyotype (`*_Basic_Karyo_TargetGenes.pdf`)**: Maps the chromosomal density of the final, resolved putative target genes (Assigned + Local) to visually identify functionally active multi-omics loci.

***

# Module 3: Expression-Aware Refinement

Physical proximity within the 3D genome is a structural prerequisite, but not a direct proxy, for active gene regulation. A chromatin loop anchoring two unexpressed loci lacks functional output. To address this, this module integrates quantitative transcriptome data to systematically eliminate transcriptionally silent physical contacts. This critical refinement distills a purely structural interactome into a high-confidence, functionally active regulatory network.

## Parameter Strategy & Core Inputs

To execute this critical functional filtration, `refine_loop_anchors_by_expression` utilizes the following key parameters:

* **`annotation_res`**: The foundational 3D annotation list generated by Module 2.
* **`expr_matrix_file` & `sample_columns`**: The quantitative expression matrix (e.g., TPM, FPKM) and the specific replicates to average for establishing a reliable baseline expression profile.
* **`threshold` & `unit_type`**: Defines the quantitative cutoff (e.g., `threshold = 1.0`, `unit_type = "TPM"`) required to consider a gene "biologically active". This allows seamless downward compatibility with various normalization methods.
* **`reclassify_by_expression`**: A transformative logical parameter (`TRUE`/`FALSE`). When enabled, mathematically silent Promoters (`P`) and Gene Bodies (`G`) are not simply discarded; they are biologically downgraded to enhancer-like regulatory elements (`eP` and `eG`). This corrects the regulatory syntax (e.g., seamlessly transitioning a functionally silent `P-P` loop into a biologically accurate `eP-P` enhancer-promoter loop).

## Example A: Standard Filtration (Strict Removal)

In this baseline scenario, we apply a strict threshold to eliminate any loops where the putative target genes are completely silent, without altering the original structural classifications.

```{r refinement_basic, message=FALSE, warning=FALSE}
res_integrated <- system.file("extdata", "analysis_results.RData", package = "looplook")

if (res_integrated != "") {
  load(res_integrated)
  res_basic <- refine_loop_anchors_by_expression(
    annotation_res = res_integrated, 
    expr_matrix_file = expr_path,
    sample_columns = c("con1", "con2"),
    threshold = 1.0,
    unit_type = "TPM",
    reclassify_by_expression = FALSE,
    out_dir = out_dir,
    project_name = "Example_Basic_Filter"
  )
}
```

## Example B: Expression-Aware Reclassification (Recommended)

This advanced mode is highly recommended. It preserves the invaluable 3D structural backbone but corrects the functional annotations. A silent promoter anchor is reclassified as an enhancer (`eP`), recognizing that while its resident gene is off, the topological locus itself is actively serving as a spatial regulatory element for a distant gene.

```{r refinement_reclassified, message=FALSE, warning=FALSE, eval = FALSE}
if (exists("res_integrated")) {
  refined_res <- refine_loop_anchors_by_expression(
    annotation_res = res_integrated,
    expr_matrix_file = expr_path,
    sample_columns = c("con1", "con2"),
    threshold = 1.0,
    unit_type = "TPM",
    reclassify_by_expression = TRUE, # Activates biological reclassification (e.g., P -> eP)
    out_dir = out_dir,
    project_name = "Example_Reclassified_Filter"
  )
}
```

## Deep Dive: Filtration Visualizations

This module automatically generates a specialized suite of visual diagnostics to quantify the impact of the transcriptome "Truth Serum" and profile the surviving functional network:

### 1. Global Filtration Profiling
*(Always Generated)*

* **Filtration Effect Dumbbell (`*_Comparison_Dumbbell.pdf`)**: Visually quantifies the "cleaning power" of the transcriptome threshold. It plots the stark numerical contrast between original structural loops and the surviving functional loops across all topological classes, highlighting the volume of un-transcribed structural noise.
* **Refined Loop Proportion Rose (`*_Rose.pdf`)**: A coxcomb chart illustrating the topological distribution (by count) of the interactome *after* expression-guided reclassification and filtration.
* **Refined Active Genes Karyotype (`*_Refined_Karyo_Active.pdf`)**: Macro-scale chromosomal ideograms mapping the physical density of the transcriptionally active loop genes that successfully passed the threshold.

### 2. 1D-to-3D Target Filtration Profiling
*(Generated only if `target_bed` was provided in Module 2)*

* **Multi-Omics Sankey Diagram (`*_Target_Sankey.pdf` / `.html`)**: The core integration diagnostic. It elegantly maps the ultimate fate of inputted 1D variants across three distinct logical flows: **Genomic Region (L1)** $\rightarrow$ **3D Loop Connection (L2)** $\rightarrow$ **Target Expression Status (L3)**. This intuitively reveals the proportion of GWAS/ChIP-seq peaks that are structurally connected to functionally active genes.
* **Refined Target Loop Donut (`*_Target_Loop_Donut.pdf`)**: Specifically visualizes the topological distribution of the functionally active loops that successfully bridge your inputted 1D targets.
* **Refined Target Genes Karyotype (`*_Refined_Karyo_TargetGenes.pdf`)**: Maps the chromosomal density of the final, strictly refined putative target genes, aiding in the visualization of transcriptionally verified multi-omics loci.

***

# Module 4: Automated Functional Profiling

Once high-confidence putative target genes are identified, `profile_target_genes` serves as a fully automated, end-to-end multi-omics analysis pipeline. It elegantly bridges 3D genomic interactions with transcriptomic data to systematically unveil the functional landscape and regulatory mechanisms of your targets.

## Parameter Strategy: A Highly Modular Pipeline

This function is exceptionally flexible. It uses a combination of routing parameters to decide exactly *which* subset of genes to analyze, and toggle parameters to turn specific downstream biological modules on or off.

### 1. Core Data Inputs
* **`annotation_res`**: The foundational annotation list (inherits seamlessly from the output of Module 2 or the refined output of Module 3).
* **`diff_file` & `lfc_col`**: Differential expression summary (e.g., from DESeq2) required for LFC profiling and GSEA.
* **`expr_matrix_file` & `metadata_file`**: Normalized expression matrix (TPM/FPKM) and sample groupings, essential for generating heatmaps and Raincloud plots.

### 2. Target Selection & Mapping Strategies
This section contains the most critical routing parameters. They dictate the fundamental biological question your downstream analysis will answer:

* **`target_source` (The Biological Scope)**: 
  * **`"targets"` *(Variant/Peak-Centric)***: Focuses **exclusively** on the putative genes regulated by your inputted 1D features (e.g., GWAS SNPs, ATAC-seq peaks). 
  * **`"loops"` *(Global Network-Centric)***: Evaluates the entire 3D interactome. It analyzes all genes anchored by the spatial loops, completely independent of your 1D peaks. 
  * **`c("loops", "targets")`**: Analyzes both scopes simultaneously, which is highly recommended for comparative profiling.
* **`target_mapping_mode`**: 
  * `"all"` *(Default)*: Accepts broad 3D target regulation (including distal enhancers looping to gene bodies).
  * `"promoter"`: Highly stringent. Strictly enforces that the 3D loop must explicitly anchor at a canonical Promoter region.
* **`include_Filled` (The Stringency Toggle)**: Logical. Controls the purity of the spatial regulation:
  * **`TRUE` *(Hybrid Mode)***: Utilizes the comprehensively merged annotation (`_Filled` columns from Module 2). It prioritizes 3D loop-derived targets but "rescues" unlooped 1D peaks by assigning them to their nearest linear gene. This ensures zero data loss and provides a complete functional overview.
  * **`FALSE` *(Pure Spatial Mode)***: Highly stringent. It strictly isolates the "pure" 3D interactome by completely dropping any 1D peaks or anchors that lack a validated 3D loop connection, isolating only structurally driven regulatory events.
* **`use_nearest_gene`**: Logical. If `TRUE`, it completely bypasses 3D loop-based assignments and strictly uses the nearest 1D linear gene, serving as a classical baseline reference (The Control).


### 3. Downstream Functional Analyses
* **`run_go`**: Executes Gene Ontology (BP) enrichment and generates Divergent Concept Networks.
* **`run_motif` & `genome_id`**: Scans proximal and distal anchors against JASPAR core motifs (requires `BSgenome`).
* **`run_ppi` & `ppi_score`**: Constructs Protein-Protein Interaction networks via the STRING database.

## Example A: Comprehensive Integrative Profiling (Recommended)

Best for complete multi-omics joint analysis (e.g., HiChIP + ATAC-seq). It comprehensively maps and profiles genes regulated by BOTH 3D distal loops and 1D direct target overlaps.

```{r profiling_scenario_A, message=FALSE, warning=FALSE}
diff_path <- system.file("extdata", "example_deg.txt", package = "looplook")
meta_path <- system.file("extdata", "example_coldata.txt", package = "looplook")

if (exists("refined_res")) {
  res_A <- profile_target_genes(
    annotation_res = res_integrated,
    diff_file = diff_path,
    lfc_col = "log2FoldChange",
    expr_matrix_file = expr_path,
    metadata_file = meta_path,
    target_source = c("loops", "targets"), # Analyzes both sources
    target_mapping_mode = "all",
    include_Filled = TRUE,
    use_nearest_gene = FALSE,
    project_name = "Scenario_A_Integrative",
    out_dir = out_dir,
    run_motif = FALSE, # Set TRUE in real analysis to scan JASPAR motifs
    run_go = FALSE, # Set TRUE in real analysis for pathway enrichment
    run_ppi = FALSE
  )
}
```

## Example B: Peak-Driven Strict Promoter Profiling (High Stringency)

This mode applies a rigorous double-filtration strategy to yield the highest confidence target gene set. 

First, setting `target_source = "targets"` ensures the engine profiles *only* the specific genes regulated by your inputted 1D peaks (e.g., your GWAS SNPs or TF binding sites), completely discarding all background spatial loops. Second, `target_mapping_mode = "promoter"` enforces topological strictness: the spatial loop bridging your peak MUST directly anchor onto the canonical promoter of the target gene (i.e., strict E-P or P-P loops), deliberately excluding looser interactions like Enhancer-to-GeneBody (E-G) connections.

```{r profiling_scenario_B, message=FALSE, warning=FALSE, eval = FALSE}
if (exists("refined_res")) {
  res_B <- profile_target_genes(
    annotation_res = res_integrated,
    diff_file = diff_path,
    lfc_col = "log2FoldChange",
    expr_matrix_file = expr_path,
    metadata_file = meta_path,
    target_source = "targets", # 1. Focus exclusively on inputted 1D peaks
    target_mapping_mode = "promoter", # 2. Require strict loop-to-promoter collision
    include_Filled = TRUE,
    use_nearest_gene = FALSE,
    project_name = "Scenario_B_StrictPromoter",
    out_dir = out_dir,
    run_go = FALSE
  )
}
```

## Example C: 1D Linear Annotation Baseline (The Control)

By setting `use_nearest_gene = TRUE`, this mode deliberately bypasses the 3D spatial topology and strictly assigns features to their nearest 1D linear genes. It is specifically designed to serve as a classical "baseline control." By comparing these results against Example A or B, you can robustly demonstrate to reviewers how much novel functional insight is gained by utilizing 3D chromatin loops over traditional nearest-gene heuristics.

```{r profiling_scenario_C, message=FALSE, warning=FALSE, eval = FALSE}
if (exists("refined_res")) {
  res_C <- profile_target_genes(
    annotation_res = res_integrated,
    diff_file = diff_path,
    lfc_col = "log2FoldChange",
    expr_matrix_file = expr_path,
    metadata_file = meta_path,
    target_source = "targets", # Focus on inputted peaks
    target_mapping_mode = "all",
    include_Filled = FALSE,
    use_nearest_gene = TRUE, # Strictly nearest 1D linear gene (The Control)
    project_name = "Scenario_C_LinearControl",
    out_dir = out_dir,
    run_go = FALSE
  )
}
```

## Deep Dive: Functional Visualizations

This module executes a massive downstream pipeline, generating an entire figure panel suitable for publication. Instead of basic data summaries, these visualizations are specifically designed to test core multi-omics hypotheses:

### 1. Expression & Topological Dynamics

* **Transcriptional Profiling (`*_LFC_Violin.pdf` & `*_Expression_Heatmap.pdf`)**: Derived from the integration of quantitative RNA-seq matrices and differential expression statistics. These visualizations assess whether the structurally annotated 3D target genes exhibit statistically significant transcriptional shifts relative to the global genomic background. This provides a quantitative framework to evaluate their potential functional relevance, offering correlative evidence of transcriptional modulation rather than definitive biological validation.
* **Topological-Transcriptional Association (`*_Connectivity_Scatter.pdf` & `*_Raincloud_*.pdf`)**: Sourced by cross-referencing 3D topological node degrees (e.g., the number of linked enhancers) with 1D expression and LFC values. This step provides a statistical framework to explore the putative "Hub Hypothesis," assessing whether an increase in spatial connectivity is positively associated with more dynamic transcriptional outputs, without asserting direct causality.

### 2. Regulatory Motif Scanning

* **Spatially Asymmetric Motif Signatures (`*_Motif_*_RankScatter.pdf` & `*_Logos.pdf`)**: Enhancer-Promoter loops are functionally asymmetric, recruiting different transcription factor (TF) families at each end. By extracting DNA sequences and scanning them against the JASPAR database *separately* for proximal (promoter) and distal (enhancer) anchors, these profiles prevent signal dilution. They decode the underlying *trans*-regulatory logic, explicitly pinpointing which distinct TF families orchestrate the two ends of the spatial chromatin architecture.

### 3. Pathway & Network Enrichment

* **Spatial Gene Set Enrichment (`*_GSEA.pdf`)**: Projects your custom, 3D-derived target gene sets onto the globally ranked differential expression landscape. Instead of relying on predefined public pathways, it treats your specific spatial network as a novel biological "gene set," statistically confirming whether your 3D interactome is a primary driver of the observed global phenotypic transition.
* **Core Effector Networks (`*_GO_Network.pdf`, `*_GO_Lollipop_*.pdf` & `*_PPI_Network_*.pdf`)**: Sourced from Gene Ontology (BP) and the STRING database. Rather than merely listing enriched terms, the Divergent Concept Networks and PPI graphs explicitly map shared biological pathways to their highest-LFC core Hub genes. This bridges the biological gap between spatial chromatin topology, transcriptomic regulation, and ultimate protein-level cooperation.

***

# Module 5: IGV-Style Track Visualization

To precisely render the local spatial interactome, this module generates a multi-tiered genomic browser view via the `plot_peaks_interactions` function. The visualization integrates the multi-omics data into three distinct, non-overlapping genomic tracks:

* **Loop Track (Top)**: Represents 3D chromatin interactions using Bezier arcs connecting spatial anchors. When `score_to_alpha = TRUE` is enabled, the opacity (alpha channel) of each arc is scaled proportionally to its quantitative interaction metric.
* **Target Track (Middle)**: Maps user-provided 1D genomic features (e.g., ChIP-seq peaks or GWAS SNPs) from the input BED file, facilitating direct visual assessment of their spatial overlap with the 3D structural anchors.
* **Gene Track (Bottom)**: Retrieves the longest canonical transcript using standard Bioconductor annotation databases, mapping exons and introns with precise strand directionality. Overlapping gene models are vertically separated to prevent label collisions.

## Parameter Strategy & Core Inputs

The function provides comprehensive control over the genomic locus, data filtering, and aesthetic mapping through the following arguments:

### 1. Data Inputs & Genomic Coordinates
* **`bedpe_file`**: Character. Path to the spatial interaction file in BEDPE format. The function utilizes the first six columns for paired anchor coordinates, and the seventh column (if present) as the quantitative interaction score.
* **`target_bed`**: Character (Optional). Path to a 1D feature file in standard BED format, rendered as the middle track for overlap assessment.
* **`chr`, `from`, `to`**: Specifies the exact genomic coordinate window for visualization. If `chr` is omitted (`NULL`), the function defaults to the most frequent chromosome present in the BEDPE file.
* **`min_score`**: Numeric (Optional). Applies a quantitative threshold to filter the displayed loops based on the score column.

### 2. Annotation Dependencies
* **`species`**: Character. Specifies the reference genome assembly. Supported inputs include `"hg38"`, `"hg19"`, `"mm10"`, or `"mm9"`. This parameter directs the function to automatically load the corresponding `TxDb` and `OrgDb` Bioconductor packages for gene track rendering.

### 3. Aesthetic Mapping & Track Configuration
* **`score_to_alpha`**: Logical. If `TRUE`, maps the quantitative interaction score to the alpha (transparency) channel of the Bezier arcs, visually differentiating interaction strengths.
* **`max_levels`**: Integer. Defines the maximum vertical stacking limit for the loop arcs to manage rendering height.
* **`base_anchor_height`**: Numeric. Sets the vertical thickness of the rectangular anchors plotted at the base of the loops.
* **Color Controls (`loop_color`, `anchor_color`, `overlap_color`, `exon_color`, `intron_color`)**: Character strings defining the specific colors (e.g., hex codes) for the structural components across all three tracks.
* **`save_file`**: Character (Optional). The file path and format extension (e.g., `".pdf"`) for exporting the final ggplot object.

```{r visual_track}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  # If 'from' and 'to' are omitted, it automatically detects the densest viewport.
  track_plot <- plot_peaks_interactions(
    bedpe_file = f1,
    target_bed = atac_path,
    chr = "chr1",
    from = 11884299,
    to = 12106581,
    species = "hg38",
    save_file = file.path(out_dir, "Locus_Track.pdf")
  )
}
```

***

# Session Info

```{r session_info, eval = TRUE}
sessionInfo()
```
