---
title: "looplook: An Integrative Framework for Chromatin Interaction Analysis and Functional Profiling"
author: 
  - name: Ying Zhang
    affiliation: Zhejiang University
    email: 12207129@zju.edu.cn
    comment: ORCID:0009-0005-9644-7062
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{looplook: An Integrative Framework for Chromatin Interaction Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 6,
  eval = FALSE  
)


```

# Introduction

**looplook** is a comprehensive R/Bioconductor toolkit designed to decipher the complex interplay between **3D chromatin architecture** (HiChIP, ChIA-PET, Hi-C) and **transcriptional landscapes**. Unlike standard tools that treat loops as static features, `looplook` implements a dynamic, biologically-aware workflow.

**Key Features & Capabilities:**

1. **3D-Guided Peak Annotation**: Leverage chromatin loops to functionally annotate "orphan" peaks (e.g., intergenic ATAC-seq peaks or GWAS variants) by physically linking them to their target genes.
2. **Expression-Aware Refinement**: Filter false-positive regulatory interactions by integrating RNA-seq data, identifying true functional Enhancer-Promoter contacts.
3. **Reproducible Consolidation**: Merge biological replicates using graph-theoretic clustering to define high-confidence interactomes.
4. **Automated Functional Profiling**: Bridge the gap between coordinates and biology via automated GO/GSEA enrichment and PPI network construction.

This vignette provides a detailed guide to the pipeline, from raw data processing to publication-quality visualization.

# Installation

```{r install}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

# Install from GitHub
devtools::install_github("your_github_username/looplook")

# Load package
library(looplook)

```

# Module 1: Data Preprocessing & Consolidation

This module focuses on standardization and reproducibility. It converts raw interaction data into robust Bioconductor objects and computes consensus sets from replicates.

## Importing Data

The `bedpe_to_gi` function is a robust importer that handles standard BEDPE formats. It automatically detects score columns (checking column 8 or 7) and normalizes anchor order (, ) to ensure compatibility with `InteractionSet`.

```{r read_data}
bedpe_file <- system.file("extdata", "example_loops_1.bedpe", package = "looplook")
gi_obj <- bedpe_to_gi(bedpe_file)

```

## Consolidating Replicates (`consolidate_chromatin_loops`)

This function addresses the challenge of merging biological replicates. Instead of simple overlaps, it employs **graph-based connected component analysis** to merge spatially proximal loops into consensus clusters.

### Parameter Details

```{r, echo=FALSE, eval=TRUE}
params <- data.frame(
  Parameter = c("files", "mode", "gap", "min_reproducible", "min_score", "blacklist_species", "target_bed_file", "out_file"),
  Description = c(
    "Character vector of file paths to BEDPE replicates (Required, n >= 2).",
    "Merge strategy: 'reproducible' (Clustering + Majority Rule), 'intersect' (Strict Overlap), or 'union' (All loops).",
    "Maximum distance (bp) to consider two anchors as overlapping (Default: 1000).",
    "Minimum number of replicates required to retain a cluster. Default is NULL (auto-calculated majority, e.g., 2/2 or 2/3).",
    "Minimum score threshold to filter input loops before merging.",
    "Genome build (e.g., 'hg38') to automatically apply ENCODE blacklist filtering.",
    "Path to a BED file. Only loops overlapping these regions are kept (e.g., Promoter-capture).",
    "Path to save the consolidated BEDPE file."
  )
)
knitr::kable(params, align = "c", caption = "Table 1: Parameters for Consolidate Chromatin Loops")

```

### Example Usage

```{r consolidate}
f1 <- system.file("extdata", "example_loops_1.bedpe", package = "looplook")
f2 <- system.file("extdata", "example_loops_2.bedpe", package = "looplook")

consensus_loops <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "reproducible",
  gap = 1000,
  min_reproducible = 2,
  out_file = "results/consensus_loops.bedpe"
)

```

# Module 2: Annotation & 3D-Guided Peak Integration

**This is a core feature of looplook.** Standard annotation tools (like ChIPseeker) annotate peaks based on the *nearest* gene. However, distal regulatory elements often regulate genes far away via looping.

The `annotate_peaks_and_loops` function uses the 3D interaction map to:

1. Classify loops (e.g., **P-P**, **E-P**) based on genomic context.
2. **Functionally annotate auxiliary peaks** (provided via `target_bed`). For example, it can determine that an intergenic GWAS SNP is physically connected to the *MYC* promoter, redefining its target gene.

### Parameter Details

```{r, echo=FALSE, eval=TRUE}
params <- data.frame(
  Parameter = c("bedpe_file", "target_bed", "species", "tss_region", "project_name", "out_dir", "color_palette", "karyo_bin_size"),
  Description = c(
    "Path to the input BEDPE file (loops).",
    "**Crucial**: Path to an auxiliary BED file (e.g., ATAC-seq, ChIP-seq, GWAS). These peaks will be annotated via loops.",
    "Genome build: 'hg38', 'hg19', 'mm10', or 'mm9'.",
    "Numeric vector defining the Promoter window around TSS (Default: c(-2000, 2000)).",
    "String prefix used for all output filenames and plot titles.",
    "Directory path to save results.",
    "Color palette for visualizations (e.g., 'Set2', 'Set3').",
    "Bin size (bp) for the genome-wide Karyotype density plot (Default: 1e5)."
  )
)
knitr::kable(params, align = "c", caption = "Table 2: Parameters for Annotation and Integration")

```

### Outputs

* **Excel Report**: Contains `Target Annotation` sheet, explicitly linking `target_bed` peaks to their 3D-connected genes via the `all_annotated_gene` column.
* **Visualizations**: Includes genomic distribution pies and connectivity bar charts.

```{r annotation}
loops_path <- "results/consensus_loops.bedpe"
atac_path <- system.file("extdata", "example_peaks.bed", package = "looplook")

anno_res <- annotate_peaks_and_loops(
  bedpe_file = loops_path,
  target_bed = atac_path,
  species = "hg38",
  out_dir = "results/annotation",
  project_name = "Study_H3K27ac"
)

```

# Module 3: Expression-Aware Refinement

Genomic annotation alone is static. A "Promoter" defined by distance to TSS might be transcriptionally silent in the cell type of interest.

The `refine_loop_anchors_by_expression` function integrates **RNA-seq (TPM)** data to dynamically refine regulatory logic. If a loop anchor is annotated as a Promoter but the gene is silent, the function can:

1. Downgrade the anchor to an **Enhancer** (assuming it might be a Poised Promoter acting as an enhancer).
2. Reclassify the loop type (e.g., **P-P**  **E-P**).

### Parameter Details

```{r, echo=FALSE, eval=TRUE}
params <- data.frame(
  Parameter = c("annotation_res", "expr_matrix_file", "sample_columns", "threshold", "downgrade_nonexpressed", "target_bed", "unit_type", "draw_original_violin", "out_dir"),
  Description = c(
    "The list object returned by 'annotate_peaks_and_loops'.",
    "Path to the expression matrix (Genes x Samples).",
    "Vector of column names/indices in the matrix to average for expression quantification.",
    "Expression cutoff (Default: 1.0). Genes < threshold are considered silent.",
    "Logical (TRUE). If TRUE, silent Promoters are reclassified as Enhancers.",
    "Optional. Re-evaluate target connectivity based on refined active genes.",
    "Label for expression units (e.g., 'TPM') for plots.",
    "Logical. Whether to draw QC violin plots of expression distribution.",
    "Output directory for refined results."
  )
)
knitr::kable(params, align = "c", caption = "Table 3: Parameters for Expression Refinement")

```

```{r refinement}
expr_file <- "data/tpm_matrix.tsv" 

refined_res <- refine_loop_anchors_by_expression(
  annotation_res = anno_res,
  expr_matrix_file = expr_file,
  sample_columns = c("Rep1", "Rep2"),
  threshold = 1.0,
  downgrade_nonexpressed = TRUE,
  out_dir = "results/refined",
  project_name = "Study_Refined"
)

```

# Module 4: Functional Profiling

The `profile_target_genes` function acts as an automated "discovery engine". It takes the refined target genes (e.g., genes regulated by E-P loops) and performs a multi-omics deep dive.

### Parameter Details

```{r, echo=FALSE, eval=TRUE}
params <- data.frame(
  Parameter = c("target_source", "diff_file", "lfc_col", "run_go", "run_ppi", "ppi_score", "stat_test", "heatmap_ntop"),
  Description = c(
    "Source of genes: 'overlap' (genes overlapping anchors) or 'cluster' (genes in the 3D hub).",
    "Path to DESeq2/edgeR result file (Must contain GeneIDs and LFC).",
    "Column name for Log2 Fold Change in the diff_file.",
    "Logical. Run Gene Ontology enrichment?",
    "Logical. Run STRING PPI network analysis?",
    "Minimum STRING interaction score (0-1000). Default: 400.",
    "Statistical test for LFC boxplots ('t.test' or 'wilcox.test').",
    "Max number of variable genes to display in the heatmap."
  )
)
knitr::kable(params, align = "c", caption = "Table 4: Parameters for Functional Profiling")

```

### Outputs

* **LFC Boxplot**: Validates if loop targets are significantly up/down-regulated compared to the background genome.
* **GSEA Plot**: Tests if loop targets are enriched in specific transcriptional programs.
* **PPI Network**: Visualizes protein-level interactions among target genes, highlighting hubs.

```{r profiling}
profile_target_genes(
  annotation_res = refined_res,
  use_filtered = TRUE,
  diff_file = "data/deseq2_results.csv",
  lfc_col = "log2FoldChange",
  tpm_file = expr_file,
  metadata_file = "data/metadata.csv",
  target_source = "cluster",
  loop_types = c("EP"), 
  run_go = TRUE,
  run_ppi = TRUE,
  out_dir = "results/profiling"
)

```

# Module 5: Visualization

`looplook` offers specialized plotting functions for both global statistics and locus-specific exploration.

## Integrative Track Plot (`plot_peaks_interactions`)

Generates a publication-ready, IGV-style static visualization. It renders chromatin loops as Bezier arcs and aligns them with gene structures and auxiliary tracks (e.g., ChIP-seq peaks).

### Parameter Details

```{r, echo=FALSE, eval=TRUE}
params <- data.frame(
  Parameter = c("bedpe_file", "target_bed", "chr/from/to", "max_levels", "base_anchor_height", "score_to_alpha", "min_score", "loop_color"),
  Description = c(
    "Path to BEDPE file.",
    "Path to auxiliary BED file (displayed as a track below loops).",
    "Genomic coordinates to visualize.",
    "Controls the height/stacking of loop arcs. Increase for dense regions.",
    "Height of the rectangular anchor blocks (Default: 0.05).",
    "Logical. If TRUE, loop transparency scales with interaction score.",
    "Minimum score to display a loop.",
    "Color for the loop arcs."
  )
)
knitr::kable(params, align = "c", caption = "Table 5: Parameters for Track Plot")

```

```{r visual_track}
plot_peaks_interactions(
  bedpe_file = loops_path,
  target_bed = atac_path,
  chr = "chr8",
  from = 127730000,
  to = 127745000,
  species = "hg38",
  save_file = "results/plots/MYC_locus.pdf"
)

```

## Contact Matrix Heatmap (`plot_contact_matrix`)

This function visualizes the aggregate density of chromatin interactions over a specific genomic region. Unlike the arc plot, it presents data as a **rotated triangular heatmap**, similar to standard Hi-C browser views. This is ideal for visualizing Topologically Associating Domains (TADs) or dense interaction hotspots.

### Parameter Details

```{r, echo=FALSE, eval=TRUE}
params <- data.frame(
  Parameter = c("gi", "region", "resolution", "shape", "palette", "max_score", "ylim_bins"),
  Description = c(
    "The GInteractions object containing loop data.",
    "String defining the genomic window (e.g., 'chr1:100000-200000').",
    "Bin size in base pairs (e.g., 5000 or 10000). Controls the 'pixel' size.",
    "Shape of the matrix pixels: 'diamond' (Classic Hi-C), 'triangle', or 'rect'.",
    "Vector of colors for the heatmap gradient.",
    "Cap value for interaction scores (to prevent outliers from washing out the plot).",
    "Limit the Y-axis height (interaction distance) to zoom in."
  )
)
knitr::kable(params, align = "c", caption = "Table 6: Parameters for Contact Matrix")

```

```{r visual_matrix}
# Define region of interest
region_str <- "chr1:9893482-13450001"

# Generate Heatmap
p_matrix <- plot_contact_matrix(
  gi = gi_obj,
  region = region_str,
  resolution = 10000,       # 10kb resolution
  shape = "diamond",        # Classic Hi-C style
  palette = c("white", "orange", "red"),
  max_score = 50            # Cap visualization at score 50
)

print(p_matrix)

```

# Session Info

Here is the output of `sessionInfo()` on the system on which this document was compiled:

```{r session_info, eval = TRUE}
sessionInfo()

```

```

```
