---
title: "Looplook: An integrative suite for target assignment and functional annotation of chromatin interactions"
author: 
  - name: Ying ZHANG
    affiliation: Zhejiang University
    email: 12207129@zju.edu.cn
    comment: ORCID:0009-0005-9644-7062
  - name: Xingze HUANG
    affiliation: Zhejiang University
    email: 22407026@zju.edu.cn
    comment: ORCID:0009-0002-9286-1344
  - name: Ye CHEN
    affiliation: Zhejiang University
    email: chenyephd@zju.edu.cn
  - name: Liang XU
    affiliation: Zhejiang University
    email: xuliang.phd@zju.edu.cn
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{looplook: An Integrative Framework for Chromatin Interaction Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
  table {
    border-collapse: collapse !important;
    border: 1px solid #ddd !important;
  }
  table th, table td {
    border: 1px solid #ddd !important;
    padding: 6px 10px !important;
  }
  table th {
    background: #f8f9fa;
    font-weight: 600;
  }
  .comparison-table-wrap {
    max-width: 100%;
    overflow-x: auto;
    overflow-y: hidden;
    margin: 30px 0;
    clear: both;
  }
  .comparison-table-wrap .horizontal-scroll {
    max-width: 100%;
    overflow-x: auto !important;
    overflow-y: hidden;
  }
  .comparison-table-wrap table {
    width: max-content;
    min-width: 100%;
    margin-bottom: 0;
  }
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 7,
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
```

***
# Introduction

**`looplook`** is an R/Bioconductor package that integrates three-dimensional (3D) chromatin architecture data — derived from HiChIP, ChIA-PET, or Hi-C — with complementary omics datasets, including transcriptomic profiles, chromatin accessibility maps, protein-DNA interaction data (ChIP-seq, CUT&Tag, CUT&RUN), and genetic variants annotated by genome-wide association studies (GWAS).

Distal regulatory elements physically contact target gene promoters through **3D chromatin looping**, enabling transcriptional regulation over distances spanning tens of kilobases to megabases along the linear genome. Conventional annotation strategies, however, assign putative regulatory elements to their **nearest neighboring genes in *cis*** — a linear approximation that frequently diverges from the underlying 3D topology. Accurate assignment of non-coding variants and orphan regulatory peaks to their **cognate target genes** therefore remains a central bottleneck in the functional interpretation of genomic regulatory landscapes.

**`looplook`** addresses this challenge by systematically prioritizing experimentally validated spatial chromatin contacts to assign target genes to thousands of regulatory elements at genome-wide scale, producing high-confidence annotations with systematic provenance tracking.

Beyond target gene assignment, **`looplook`** functions as a **standalone platform** for chromatin loop analysis. Even in the absence of auxiliary omics data, the package enables systematic annotation of the 3D interactome, classification of spatial interaction topologies (e.g., Enhancer–Promoter, Promoter–Promoter), and quantification of node connectivity to identify **dense regulatory hubs** and **enhancer cliques** — candidate regulatory domains that may drive cell-type-specific transcriptional programs.

## Key Features

1. **3D-Guided Multi-Omics Integration**: Maps auxiliary genomic features (GWAS SNPs, ChIP-seq peaks, ATAC-seq regions) to target genes via 3D chromatin contacts. A structured fallback mechanism prioritizes loop-mediated target genes and supplements them with nearest-gene assignments only when no spatial interaction is detected, ensuring both 3D specificity and complete coverage.

2. **Systematic Loop Annotation and Hub Detection**: Classifies spatial interaction types (E–P, P–P, E–E, and their expression- and chromatin-refined variants) and computes node degree distributions to identify candidate 3D regulatory hubs and enhancer cliques.

3. **Expression-Aware Refinement**: Integrates quantitative transcriptomic data to annotate each loop with expression-informed functional status while preserving all structural contacts. Anchors associated with transcriptionally silent reference promoters are reclassified as enhancer-like elements (eP, eG), producing an expression-refined regulatory network without discarding structural evidence. A dataset-adaptive quantile mode (`threshold_mode = "quantile"`) accommodates varying sequencing depths.

4. **Chromatin-Guided Anchor Reclassification**: Validates and systematically reclassifies loop anchors using orthogonal chromatin mark data (ChIP-seq, CUT&Tag, ATAC-seq). Anchors are scored against ENCODE active-enhancer criteria across five confidence levels (gold\_standard to uncertain), and anchor types are updated based on mark combinations — for instance, an eP anchor with H3K4me3+ H3K27me3− promoter-like chromatin is reverted to P, while a P anchor with H3K4me1+ H3K4me3+ dual marks is upgraded to dual-function. Reclassification is fully auditable through per-anchor chromatin state, confidence, and before/after type columns.

5. **Replicate Consolidation and Multi-Source Consensus**: Employs graph-theoretic connected-component clustering to harmonize biological replicates and suppress technical noise. The framework supports multi-source consensus integration — for instance, intersecting HiChIP datasets targeting H3K27ac and RNA Polymerase II — to construct a unified, high-confidence 3D interactome supported by orthogonal experimental evidence.

6. **Automated Functional Profiling**: Delivers an end-to-end analytical pipeline generating publication-ready visualizations and functional interpretations, including network connectivity and expression dynamics, transcription factor motif enrichment (SeqLogos), and pathway association networks (Cnetplots).

## Comparison with Existing Tools

`looplook` bridges the gap between conventional 1D linear annotation and specialized 3D chromatin interaction tools, providing a unified environment for target assignment, expression-aware refinement, consensus consolidation, and functional inference. Table \@ref(tab:comparison-table) summarizes the core capabilities of `looplook` relative to existing tools in the ecosystem.

<div class="comparison-table-wrap">

```{r comparison-table, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
# 1. Define symbols
yes <- "<span style='color: #2E8B57;'>&#10004;</span>"
no <- "<span style='color: #B0BEC5;'>&#10008;</span>"
part <- "<span style='color: #FF8F00;'>Partial</span>"

# 2. Build comparison table
comp <- data.frame(
  "Feature" = c(
    "Primary annotation scheme", "User-input 3D interactome", "User-input omics data",
    "Graph-based topological consensus", "Expression-guided refinement",
    "Chromatin mark validation", "Topological reclassification",
    "Multi-hop network diffusion", "Smart linear fallback", "Functional inference",
    "Multi-track visualization", "Local execution", "Ecosystem"
  ),
  ChIPseeker = c("1D", no, yes, no, no, no, no, no, no, no, no, yes, "R"),
  GREAT = c("1D", no, yes, no, no, no, no, no, no, yes, no, no, "Web"),
  GenomicInteractions = c("3D", yes, no, no, no, no, no, no, no, no, no, yes, "R"),
  FUMA = c("3D", no, no, no, part, no, no, no, no, no, no, no, "Web"),
  "ABC Model" = c("3D", yes, yes, no, no, no, no, no, no, no, no, yes, "Python"),
  looplook = c("1D & 3D", yes, yes, yes, yes, yes, yes, yes, yes, yes, yes, yes, "R"),
  check.names = FALSE, stringsAsFactors = FALSE
)

# 3. Render with or without kableExtra
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  knitr::kable(
    comp,
    format = "html",
    escape = FALSE,
    align = c("l", "c", "c", "c", "c", "c", "c"),
    caption = "Comparison of core features between looplook and existing genomic annotation tools."
  )
} else {
  knitr::kable(
    comp,
    format = "html",
    escape = FALSE,
    align = c("l", "c", "c", "c", "c", "c", "c"),
    caption = "Comparison of core features between looplook and existing genomic annotation tools."
  ) %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE,
      position = "left",
      font_size = 12
    ) %>%
    kableExtra::row_spec(
      0,
      bold = TRUE,
      color = "black",
      background = "#F8F9FA",
      extra_css = "border: 1px solid #DDDDDD;"
    ) %>%
    kableExtra::column_spec(1:7, extra_css = "border: 1px solid #DDDDDD; padding: 8px;")
}
```

</div>

<small style="color: #666; display: block; margin-top: -10px; margin-bottom: 20px;">
<em>Notes:</em> “1D” = linear proximity-based mapping; “3D” = mapping via spatial chromatin topology; “<span style='color: #2E8B57;'>&#10004;</span>” = native support; “<span style='color: #B0BEC5;'>&#10008;</span>” = not supported; “<span style='color: #FF8F00;'>Partial</span>” = partial support (FUMA relies on precompiled data rather than user-provided input).
</small>

`looplook` provides both 1D and 3D target annotation in a unified R workflow. Its three-stage refinement strategy — expression-aware reclassification of silent regulatory elements (P → eP, G → eG), orthogonal chromatin-mark validation against ENCODE enhancer criteria, and systematic anchor-type reclassification based on mark combinations — distinguishes structural chromatin contacts from functionally active regulatory events. Graph-based replicate consolidation harmonizes loop calls across biological replicates without over-merging, while per-Anchor provenance tracking ensures full auditability of every reclassification decision.

***

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("zying106/looplook")
```

```{r load_package}
library(looplook)
# Create a temporary directory to store all output plots and Excel files
out_dir <- tempdir()
```

***

# Module 1: Data Consolidation & Preprocessing

Chromatin interaction datasets from individual biological replicates invariably contain technical variability and stochastic noise. The **`consolidate_chromatin_loops`** function serves as the foundational data-integration module, merging multiple replicates into a standardized, unified 3D interaction coordinate framework.

The module supports **multi-source consensus** integration across independent experimental designs — for instance, HiChIP assays targeting H3K27ac and RNA Polymerase II. By identifying shared spatial interactions across heterogeneous inputs, the framework enables **orthogonal validation** that mitigates assay-specific biases and technical artifacts. Consensus loops supported by multiple independent lines of evidence provide a robust foundation for downstream annotation and functional interpretation.

### Parameter Strategy

* **`mode`**: Defines the overarching merging algorithm.
    * `"consensus"` *(Recommended)*: Implements graph-based connected component analysis to cluster spatially proximal anchors across samples. Only retains clusters detected in >= `min_consensus` biological replicates.
    * `"intersect"`: Enforces strict reference-based filtering, retaining loops that show full genomic overlap with the reference file (File 1). 
    * `"union"`: Retains all chromatin interactions across the entire cohort, ideal for exploratory pan-tissue analyses.
* **`min_consensus`**: When using `"consensus"` mode, this parameter defines the minimum number of biological replicates in which a loop cluster must be detected to be retained in the final dataset. If set to `NULL`, the algorithm automatically computes the threshold: 100% for 2–3 replicates, and ≥75% (ceiling(0.75 × N)) for N ≥ 4 (e.g., 3 for N=4, 4 for N=5, 6 for N=8).
* **`gap`**: Defines the maximum allowable spatial distance (in base pairs) between loop anchors for them to be considered as part of the same physical cluster (default: `1000`). When `quiet = FALSE` (default), the function prints a pre-clustering gap diagnosis and a post-clustering cluster-span diagnosis, including data-driven gap recommendations and chaining risk assessment. If loop anchors are narrow (e.g., TF HiChIP, 200–400 bp), the diagnosis may suggest reducing `gap` to avoid over-merging independent regulatory events.
    * **How to choose `gap`**: As a rule of thumb, `gap` should be comparable to (not much larger than) the median loop anchor width in your data. For histone-mark HiChIP (H3K27ac, ~1–3 kb anchors), `1000` is appropriate. For TF HiChIP or ChIA-PET (100–500 bp anchors), consider `100–200`. For Hi-C bins (5–50 kb), `1000` is conservative. The automatic diagnosis will guide you if the chosen `gap` is inappropriate for your data.
* **`min_raw_score` vs. `min_score` (The Dual-Filter)**:
    * `min_raw_score` acts as a pre-filter applied to individual BEDPE files before clustering (e.g., removing singleton noise where PET count < 2) to help reduce computational memory overhead.
    * `min_score` serves as a post-filter applied to the final merged chromatin interactome to ensure high-confidence interactions. In `"consensus"` and `"union"` modes, this representative score is computed in a replicate-balanced manner: loop scores are averaged within each replicate first, then averaged across replicates.
* **`blacklist_species`**: Automatically excludes chromatin loops overlapping with high-variance, artifact-prone genomic regions (e.g., centromeres, telomeres). Accepts a species name (`"hg38"`, `"hg19"`, `"mm10"`, `"mm9"`) to use the bundled ENCODE blacklist, or a path to a custom blacklist BED file.
* **`region_of_interest`**: Accepts an auxiliary BED file (e.g., a specific disease-associated locus or ChIP-seq peak set) to capture only loops with physical connectivity to the target genomic region of interest (ROI).
* **`roi_mode`**: Controls how strictly loops must overlap the ROI:
    * `"any"` *(Default)*: Either anchor may overlap the ROI. Suitable for promoter-centric queries (e.g., find all distal interactions involving a specific gene locus).
    * `"both"`: Both loop anchors must overlap the ROI. Suitable for TAD confinement, domain-internal interaction queries, or when filtering with a ChIP-seq peak set (e.g., keep only loops where both anchors fall in H3K27ac-marked regions).
* **`score_col`**: Integer column index for the interaction score (e.g., PET count) when BEDPE files follow a non-standard column layout. If `NULL` (default), columns 8 and 7 are auto-detected by checking the proportion of numeric values.

\strong{Score direction: All score-based filtering assumes higher scores = better interactions.} If your BEDPE score column contains p-values (values in [0, 1]), convert them to `-log10(p-value)` before using `min_raw_score` or `min_score`. The function will warn if it auto-detects a p-value-like column, but explicit conversion is the safest approach.
* **`quiet`**: Logical. If `TRUE`, suppress progress messages while preserving warnings (default: `FALSE`).

The exported BEDPE file (when `write_output = TRUE` and `out_file` is set) contains 10 columns: the standard BEDPE fields (chr1, start1, end1, chr2, start2, end2, name, score) plus two metadata columns:
- **`n_members`** (column 9): Number of raw loops merged into this consensus entry, giving a sense of cluster size. In intersect mode this is `1` (no merging, each retained loop stands alone); in consensus/union modes it reflects the number of loops clustered.
- **`n_reps`** (column 10): Number of input files that support this entry, useful for filtering by replication evidence.

## Example 1: Building a Global Consensus Interactome

This represents the standard workflow for generating a consolidated whole-genome 3D chromatin interaction dataset via integration of biological replicates and removal of blacklist-associated artifacts.

```{r consolidate_global, message=FALSE, warning=FALSE}
# Locate example BEDPE replicates
f1 <- system.file("extdata", "example_loops_1.bedpe", package = "looplook")
f2 <- system.file("extdata", "example_loops_2.bedpe", package = "looplook")
global_out <- file.path(out_dir, "consensus_loops_global.bedpe")

# Execute consensus merging with strict Quality Control
consensus_global <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2, # Pre-filter sequencing noise
  blacklist_species = "hg38", # Apply artifact blacklist
  out_file = global_out
)
```

## Example 2: Feature-Driven Loop Filtering

In most scenarios, researchers are interested in 3D chromatin interactions associated with specific biological features, such as active enhancers (marked by H3K27ac) or particular transcription factors. 

By providing a BED file to `region_of_interest`, the tool enables efficient "on-the-fly" capture, retaining only those loops whose anchors overlap with the specified functional regions.

```{r consolidate_targeted, message=FALSE, warning=FALSE}
# Locate a BED file containing H3K27ac ChIP-seq peaks
h3k27ac_peaks <- system.file("extdata", "example_k27ac_peaks.bed", package = "looplook")
targeted_out <- file.path(out_dir, "H3K27ac_anchored_loops.bedpe")

# Execute consensus merging, strictly capturing H3K27ac-associated loops
consensus_targeted <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2,
  region_of_interest = h3k27ac_peaks,
  out_file = targeted_out
)
```

## Example 3: Multi-source Consensus for Orthogonal Validation

In integrative multi-omics analyses, orthogonal validation across independent functional assays is critical for identifying robust 3D regulatory interactions. For instance, enhancer-promoter (E-P) loops are typically supported by the spatial co-localization of H3K27ac (marking active chromatin) and RNA polymerase II (indicative of transcriptional machinery engagement). 
By utilizing the `"consensus"` mode to intersect independent loop datasets, `looplook` facilitates the identification of chromatin interactions supported by multi-source evidence. This conservative intersection strategy may reduce assay-specific technical artifacts, yielding a refined set of loops supported by multiple assays.


```{r consolidate_multi_modal, message=FALSE, warning=FALSE}
loop_k27ac <- system.file("extdata", "example_loops_H3K27ac.bedpe", package = "looplook")
loop_pol2 <- system.file("extdata", "example_loops_pol2.bedpe", package = "looplook")
dual_functional_out <- file.path(out_dir, "Dual_Functional_Consensus.bedpe")

consensus_dual <- consolidate_chromatin_loops(
  files = c(loop_k27ac, loop_pol2),
  mode = "consensus",
  gap = 1000,
  min_raw_score = 2,
  out_file = dual_functional_out
)
```

***

### Module 2: 3D-Guided Annotation & Mapping

**`annotate_peaks_and_loops`** serves as the core mapping engine of **`looplook`**. It classifies the topological architecture of the 3D interactome — assigning each loop anchor a positional type (P, E, G) and each loop a structural class (E–P, P–P, etc.) — and, when a `target_bed` file is supplied, traces non-coding regulatory features (GWAS loci, eQTLs, ChIP-seq peaks) to their putative target genes through 3D chromatin contacts.

To resolve mapping ambiguities within densely populated gene loci, where a single loop anchor overlaps multiple candidate genes, the engine implements a rigorous three-step hierarchical annotation pipeline:

1. **Functional Biotype Prioritization**: Selects the highest-priority candidates by functional class: Protein Coding > small-ncRNA (miRNA, snoRNA, snRNA, rRNA, scaRNA) > Antisense > lncRNA/ncRNA > Pseudogene. This default hierarchy can be customised via the `biotype_order` parameter in `resolve_gene_conflicts()`. The five biotype keywords are `"protein"` (protein-coding), `"small_ncRNA"` (miRNA, snoRNA, snRNA, rRNA, scaRNA), `"antisense"`, `"lncRNA"` (lncRNA and other ncRNA), and `"pseudogene"`. The order of listed keywords determines priority (first = highest); unlisted categories keep their default relative order and are appended after the listed ones. For example, `biotype_order = c("lncRNA", "protein")` elevates lncRNAs above protein-coding genes while keeping the remaining three categories (small_ncRNA > antisense > pseudogene) in their default order.
2. **Expression Filter**: Within the selected biotype tier, excludes transcriptionally silent genes (expression < `min_expr`) if at least one candidate in the tier is expressed. If no gene in that tier is expressed, all are retained to avoid losing correct annotations due to low expression.
3. **Co-Dominant Expression Tiebreaker**: Among remaining candidates, retains all genes with expression >= `co_dominance_ratio` x the group maximum. The default ratio of 0.1 (one order of magnitude; ~3.3-fold in log2 space) preserves functionally redundant candidates (e.g., bidirectional promoter pairs, typically co-expressed within 2–3 fold) rather than forcing a single-gene assignment.

### Parameter Strategy & Core Inputs

This multi-omic integration relies on several key parameters:

* **`bedpe_file`**: The 3D structural framework (e.g., the high-confidence consensus interactome generated in Module 1).
* **`target_bed`**: The auxiliary genomic features of interest (e.g., a BED file of GWAS SNPs, ATAC-seq peaks, or ChIP-seq peaks) awaiting spatial target assignment.
* **`expr_matrix_file` & `sample_columns`**: Optional but strongly recommended. An expression matrix (e.g., RNA-seq) enables the Expression Pre-filter and Tiebreaker logic, which can help reduce false-positive gene assignments in dense genomic loci.
* **`min_expr`**: Minimum expression value for a gene to be considered "active" during anchor-level conflict resolution (default: `0`). At the default, any non-zero expression (TPM > 0) qualifies; truly undetected genes (TPM = 0) are excluded. When set to `1` or higher, genes with expression `>= min_expr` are considered active. Note this is distinct from Module 3's **`threshold`** parameter, which controls promoter reclassification during expression-aware refinement.
* **`neighbor_hop`**: An advanced topological parameter for network traversal. Note: target gene assignment searches one additional graph step (`neighbor_hop + 1`) to capture genes at the opposite anchor of directly connected loops, while loop-level topology uses the specified hop count.
    * `0` *(Default)*: Loop topology restricted to direct contacts; target genes searched within 1-hop.
    * `1` *(Hub Mode)*: Loop topology includes secondary contacts; target genes searched within 2-hop.
* **`anchor_gap`** (integer, default `-1`): Search radius between a target peak and a loop anchor. `-1` (default): strict physical overlap (GenomicRanges default — peak and anchor must share at least 1 bp). `0`: adjacent/touching intervals also count. `>0`: explicit gap tolerance in bp. When `>= 0`, the result includes both physically overlapping pairs AND proximity-only pairs (within gap but no actual overlap). Use `anchor_min_overlap > 1` to require physical overlap among these candidates.
* **`anchor_min_overlap`** (integer, default `1`): After candidates are found via `anchor_gap`, how many bp of actual physical overlap are required? `1` (default): any touch counts. Increase to `10–100` to filter out spurious boundary overlaps. Applied after `anchor_gap`.
* **`anchor_min_frac`** (numeric, 0–1, default `0`): After the first two filters, what fraction of the *peak* width must physically overlap the anchor? `0` accepts any fraction. Set to `0.1–0.5` when peaks are broad (H3K27ac domains, 2–5 kb). Not recommended for point features (SNPs). Applied last.

The three parameters act as a cascade: (1) search radius, (2) minimum overlap bp, (3) fraction of peak. They do not conflict; each narrows the candidate set further.

* **`tss_region`**: Defines the genomic window surrounding the transcription start site (TSS) used to define promoter regions (default: `c(-2000, 2000)` bp).
* **`conflict_strategy`**: Controls the order of biotype and expression filtering when multiple genes overlap an anchor. `"biotype_first"` (default) selects the best biotype tier first, then applies expression filtering within that tier — a silent protein-coding gene is preferred over a highly expressed lncRNA. `"expression_first"` applies expression filtering across all biotypes before selecting the best tier.
* **`hub_percentile`**: Node-degree quantile threshold for identifying highly connected regulatory elements (default: `0.95`, i.e. top 5%).
* **`species`**: Genome assembly (`"hg38"`, `"hg19"`, `"mm10"`, or `"mm9"`). Auto-resolves the corresponding `TxDb` and `OrgDb` annotation packages. **For other species**, pass `txdb` and `org_db` as objects or package name strings directly (e.g., `txdb = TxDb.Dmelanogaster.UCSC.dm6.ensGene`, `org_db = "org.Dm.eg.db"`). When both `txdb` and `org_db` are provided, `species` is only used for karyotype chromosome ideograms and can be left as the default.
* **`color_palette`**: RColorBrewer palette for loop-type colour assignments in annotation plots (default: `"Set2"`).
* **`karyo_bin_size`**: Genomic bin width in base pairs for karyotype heatmaps (default: `1e5`).

#### Suggested Anchor Overlap Settings (for reference)

| Experimental design | `anchor_gap` | `anchor_min_overlap` | `anchor_min_frac` |
|---------------------|:---:|:---:|:---:|
| Same-experiment HiChIP / ChIA-PET (default) | `-1` | `1` | `0` |
| Cross-experiment (ATAC-seq peaks × HiChIP loops) | `200–500` | `10` | `0` |
| Broad histone-mark peaks (H3K27ac, 2–5 kb) | `0` | `100` | `0.1` |
| Super-enhancers / wide domains (20–80 kb) | `0` | `500` | `0.05` |
| Point features (GWAS SNPs, eQTLs) | `0` | `1` | `0` |
| High-confidence only (stringent) | `0` | `100` | `0.5` |

## Example A: Integrative Analysis (Loops + Genomic Features + RNA-seq)

This represents the comprehensive “3D Spatial Bridge” scenario. It links external genomic feature (peaks) to 3D‑inferred target genes, while simultaneously resolving ambiguous assignments in dense genomic regions by weighing gene expression.

```{r annotation_integrated, message=FALSE, warning=FALSE, results='hide'}
# Locate auxiliary genomic features and RNA-seq expression matrix
expr_path <- system.file("extdata", "example_tpm.txt", package = "looplook")
atac_path <- system.file("extdata", "example_peaks.bed", package = "looplook")
if (requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  res_integrated <- annotate_peaks_and_loops(
    bedpe_file = global_out, # Uses the global consensus network from Module 1
    target_bed = atac_path, # External genomic features
    expr_matrix_file = expr_path, # Activates the rigorous 3-step tiebreaker
    sample_columns = c("con1", "con2"),
    species = "hg38",
    neighbor_hop = 0, # Focus on direct physical contacts
    hub_percentile = 0.95, # Top 5% nodes defined as hubs
    # Peak-anchor overlap: same-experiment defaults (strict physical overlap)
    anchor_gap = -1L,           # GenomicRanges default: strict overlap
    anchor_min_overlap = 1L,    # any 1 bp touch counts
    anchor_min_frac = 0,        # any fraction of peak is accepted
    out_dir = out_dir,
    project_name = "Example_HiChIP_Integrative"
  )
}
```

## Example B: Intrinsic Loop Profiling (Loops ONLY)

For users focused exclusively on 3D interactome architecture without auxiliary data related to genomic features, **`looplook`** functions as a standalone platform for loop profiling and structural hub identification.


```{r annotation_loops_only, message=FALSE, warning=FALSE, results='hide'}
if (requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  res_loops_only <- annotate_peaks_and_loops(
    bedpe_file = global_out,
    target_bed = NULL, # Omit genomic features entirely
    expr_matrix_file = expr_path, # Resolves multi-gene conflicts at loop anchors
    sample_columns = c("con1", "con2"),
    species = "hg38",
    neighbor_hop = 0,
    out_dir = out_dir,
    project_name = "Example_Loops_Only"
  )
}
```

### Deep Dive: Output Data Dictionary

The function exports a comprehensive Excel workbook (`*_Basic_Results.xlsx`) containing multi-layered annotations. Interpretation of spatial regulatory logic depends critically on the hierarchical structure of its columns:

#### 1. `target_annotation` (The Genomic Feature-to-3D Mapping Result)
*(Generated only when `target_bed` is provided)*

* **`annotation`**: Genomic context of the input peaks (e.g., "Distal Intergenic", "Promoter").
* **`Linked_Loop_IDs`**: Identifiers of 3D loops that overlap the peak.
* **`All_Loop_Connected_Genes`**: All genes (including promoters and gene bodies) topologically linked to the peak via 3D looping.
* **`Regulated_promoter_genes`**: A stringent promoter-target set supported by loop-anchor context. A promoter gene may be local to the overlapped anchor or located on the opposite/interacting anchor; the evidence type is recorded separately.
* **`Regulated_promoter_Evidence`**: Semicolon-delimited evidence labels for `Regulated_promoter_genes`, such as `local_promoter_overlap`, `direct_opposite_promoter`, or `expanded_promoter_loop`.
* **`Assigned_Target_Genes`**: The historical promoter-first 3D assignment. It prefers promoter genes reachable through the loop neighborhood and falls back to gene-body loop genes only when no promoter target is available.
* **`*_Filled` Columns (The "Smart Fallback" Logic)**: Columns suffixed with `_Filled` (e.g., `Assigned_Target_Genes_Filled`, `Regulated_promoter_genes_Filled`) provide complete, gapless annotations. They use strict 3D assignments first and add the nearest linear gene (`SYMBOL`/`geneId`) only when the corresponding strict column is empty.
* **`Regulated_promoter_Fallback_Evidence`**: Records why `Regulated_promoter_genes_Filled` used a linear fallback (`local_promoter`, `local_gene_body`, `linear_nearest`, or `linear_fallback`); it is `none` when strict loop-derived promoter targets exist.
* **`target_gene_links`**: A long-format provenance table. Each row is one peak-gene link with `source`, `evidence`, `anchor_role`, `loop_ID`, fallback status, and membership flags for the summary columns. In the refined workbook, `Filtered Target Gene Links` keeps only links still used by the refined target columns and adds `Mean_Expression` plus `Passes_Expression_Filter`.
* **`SYMBOL`**: The gene symbol resolved by the biotype-aware conflict resolution logic. May be a single gene or semicolon-delimited when multiple co-dominant candidates are retained.
* **`loop_type`**: The dominant loop topology among the 3D interactions that overlap the feature.
* **`All_Loop_Connected_Genes_Filled`**: Lineage column that preserves the original loop-connected gene set and fills empty entries with the nearest linear gene fallback.

\newpage

##### 1a. Evidence label reference

**`Regulated_promoter_Evidence`** — provenance of `Regulated_promoter_genes` (from 3D loop anchors):

| Value | Meaning |
|-------|---------|
| `local_promoter_overlap` | The peak overlaps the promoter of the anchor gene |
| `direct_opposite_promoter` | The opposite anchor of the linked loop is a promoter |
| `gene_body_context` | The anchor overlaps a gene body (not a promoter) |
| `expanded_promoter_loop` | The promoter gene was reached via ego-network expansion (`neighbor_hop`) |
| `none` | No loop-derived promoter evidence found |

Multiple evidence types may be combined in one cell (semicolon-delimited).

**`Regulated_promoter_Fallback_Evidence`** — provenance when `Regulated_promoter_genes_Filled` used a linear fallback:

| Value | Meaning |
|-------|---------|
| `none` | Strict loop-derived promoter targets exist; no fallback needed |
| `local_promoter` | Fallback gene overlaps a promoter region near the peak |
| `local_gene_body` | Fallback gene overlaps an exon/intron/UTR region |
| `linear_nearest` | Fallback gene is the nearest gene by linear distance |
| `linear_fallback` | Fallback from annotation column with no location context |

**`target_gene_links`** — per-link label reference:

| Column | Possible values |
|--------|-----------------|
| `source` | `loop_anchor`, `linear_annotation` |
| `gene_role` | `promoter`, `gene_body`, `linear_annotation` |
| `evidence` | `local_promoter_overlap`, `direct_opposite_promoter`, `gene_body_context`, `expanded_promoter_loop`, `linear_annotation`, `linear_fallback` |
| `anchor_role` | `local_anchor`, `opposite_anchor`, `expanded_anchor`, `linear_annotation` |
| `used_as_fallback` | `TRUE`, `FALSE` |

Membership flags (`in_regulated_promoter`, `in_assigned_target`, `in_all_loop_connected`, `in_regulated_promoter_filled`, `in_assigned_target_filled`) indicate which target annotation column(s) contain the gene; refined links additionally carry `Mean_Expression` and `Passes_Expression_Filter`.

#### 2. `loop_annotation` (3D Network Architecture)

* **`loop_type`**: Topological classification (e.g., E-P, P-P, E-E) determined by the biotype-aware logic.
* **`All_Anchor_Genes`**: All genes located linearly within the left and right anchors of a given loop.
* **`Putative_Target_Genes`**: Biologically resolved target genes. For an E-P loop, this field will report only the gene at the Promoter anchor.
* **`Cluster_All_Genes`**: All genes within the broader 3D cluster or clique in which the loop resides, reflecting the extended topological neighborhood.
* **Coordinate columns** (`chr1`/`start1`/`end1`, `chr2`/`start2`/`end2`): 1-based genomic coordinates of the two loop anchors.
* **`anchor1_gene` / `anchor2_gene`**: Gene symbols associated with each anchor after biotype-aware conflict resolution.
* **`anchor1_type` / `anchor2_type`**: Positional classification of each anchor relative to gene annotations (`P` = promoter, `E` = distal/intergenic — positional, not functional enhancer, `G` = gene body; `eP`/`eG` after refinement indicate expression-filtered silent P/G anchors).
* **`loop_ID`**: Unique loop identifier; `cluster_id`: connected-component membership.

#### 3. `anchor_loci_annotation` (Non-redundant Spatial Anchor Footprints)

* **`Cluster_Locus_Genes`**: All genes from loop anchors within a spatially interconnected component (defined by network-connected components), representing the full gene set associated with the deduplicated anchor loci.
* **Backward compatibility**: `anchor_annotation` is retained as an alias of `anchor_loci_annotation` for older scripts.

#### 4. `promoter_centric_stats` & `distal_element_stats` (Topological Hub Detection)

These metrics quantify interactome connectivity to identify candidate regulatory hubs from complementary perspectives:

**A. Gene-Centric (`promoter_centric_stats`)**

- **`Total_Loops` / `n_Linked_Promoters` / `n_Linked_Distal`**: Deconstructs a gene's 3D connectome. High `n_Linked_Distal` suggests potential regulation by enhancer clusters, whereas high `n_Linked_Promoters` may indicate possible participation in multi-gene transcription.
- **`Dominant_Interaction`**: The predominant topological class governing the gene (e.g., E-P vs. P-P). When multiple loop types tie for the highest frequency, the first in alphabetical order is selected (e.g., E-P before P-P). This tie-breaking is deterministic but arbitrary; inspect raw counts when ties are likely (small datasets).
- **`Is_High_Connectivity_Gene`** *(and Distal variant)*: Binary flags defined by the `hub_percentile` to prioritize putative 3D regulatory hubs.

**B. Enhancer-Centric (`distal_element_stats`)**

- **`Target_Genes`**: Promoters physically contacted by the distal anchor.
- **`n_Linked_Promoters`**: Facilitates identification of candidate master enhancers that contact multiple distinct genes.
- **`Is_High_Connectivity_Distal_Element`**: Flags highly connected non-coding elements that may act as 3D hubs.

After expression-aware refinement (`*_Refined_Results.xlsx`), the promoter-centric stats additionally contain:

- **`Is_Active_Gene`**: `"Yes"` when the gene's mean expression meets or exceeds the `threshold`; `"No"` otherwise. This column distinguishes structurally connected genes from those that are also transcriptionally active in the current sample context.
- Hub classification columns (`Is_High_Connectivity_Gene`, `Is_High_Distal_Connectivity_Gene`) reflect the refined network topology, independent of expression status.

### Deep Dive: Output Visualizations {.tabset}

All plots are returned as objects in `res_integrated$plots$`, enabling direct inspection and customisation with standard **ggplot2** layers without re-running the pipeline.

#### 1. Global 3D Network Profiling
*(Always Generated)*

| Plot key | Description |
|---|---|
| `Basic_Donut` | Loop-type frequency (donut chart) |
| `Basic_Circular` | Unique target genes per loop type (circular bar) |
| `Basic_Flower` | Shared vs. unique genes across loop types |
| `Karyo_Anchors` | Genome-wide anchor density heatmap |
| `Karyo_LoopGenes` | Genome-wide loop-associated gene density |
| `Anchor_Genomic_Distribution` | Genomic context of anchors (pie) |
| `Target_Genomic_Distribution` | Genomic context of input features (pie) |

#### 2. Genomic Feature-to-3D Target Profiling
*(Requires `target_bed`)*

| Plot key | Description |
|---|---|
| `Target_Rose` | Loop types linked to input features (coxcomb) |
| `Target_Loop_Genomic_Distribution` | Genomic context of loop-connected features (pie) |
| `Karyo_TargetGenes` | Chromosomal density of assigned target genes |

#### Customising Plots

All ggplot-based plots accept standard **ggplot2** layers for recolouring, theme adjustment, or annotation:

```{r plot-customise, eval=exists("res_integrated") && !is.null(res_integrated$plots$Basic_Donut)}
# Access a stored plot
p <- res_integrated$plots$Basic_Donut

# Extract data and original label mapping
donut_data <- p$data
labels_vec <- setNames(donut_data$legend_label, donut_data$loop_type)

# Apply a custom colour palette
p + ggplot2::scale_fill_manual(
  values = RColorBrewer::brewer.pal(length(labels_vec), "Dark2"),
  labels = labels_vec
) + ggplot2::theme(legend.position = "bottom")
```

To save any plot to disk, use `ggplot2::ggsave()`:

```{r plot-save, eval=exists("res_integrated")}
ggplot2::ggsave(
  filename = file.path(out_dir, "my_custom_donut.pdf"),
  plot = res_integrated$plots$Basic_Donut,
  width = 8, height = 6
)
```

For `looplook_karyo` objects (karyotype heatmaps), use `print()` to display:

```{r plot-karyo, eval=exists("res_integrated")}
print(res_integrated$plots$Karyo_Anchors) # renders via grid or opens in browser
```

***

# Module 3: Refinement & Validation

Physical proximity detected by chromatin conformation assays is a structural prerequisite for regulatory interaction, but it does not directly indicate active transcriptional engagement. This module integrates quantitative transcriptomic data to annotate each loop with expression-informed functional status. All structural loops are preserved; the pipeline reclassifies transcriptionally silent anchors (P → eP, G → eG), designates the high-confidence functional subset via `Retained_In_Functional_Network`, and records the rationale for each decision in `Refinement_Action`.


### Parameter Strategy & Core Inputs

The `refine_loop_anchors_by_expression` function implements expression-aware refinement via the following key parameters:

- **`annotation_res`**: Foundational 3D annotation output generated by Module 2.
- **`expr_matrix_file` & `sample_columns`**: Quantitative expression matrix (e.g., TPM, FPKM) and the specific replicates to average, establishing a robust baseline expression profile.
- **`threshold` & `unit_type`**: Defines the quantitative expression cutoff (e.g., `threshold = 1`, `unit_type = "TPM"`); genes with expression >= `threshold` are considered active. This parameter enables downward compatibility with various normalization methods.
- **`threshold_mode`** (new): Controls how `threshold` is interpreted. `"absolute"` (default): `threshold` is a direct expression cutoff (e.g., TPM >= 1). `"quantile"`: `threshold` is interpreted as a quantile of the expression distribution (e.g., `0.75` selects the top 25% most highly expressed genes). Quantile mode is dataset-adaptive and robust across experiments with different sequencing depths.
- **`reclassify_by_expression`**: A transformative logical parameter (`TRUE`/`FALSE`). When enabled, transcriptionally silent Promoters (`P`) and Gene Bodies (`G`) are not simply discarded; instead, they are reclassified as enhancer-like regulatory elements (`eP`, `eG`). This correction refines the regulatory topology (e.g., reclassifying a functionally silent `P-P` loop into a curated `eP-P` loop). **Important:** `eP`/`eG` labels indicate transcriptional silence at the reference gene — they do **not** constitute evidence of enhancer activity. Orthogonal chromatin data (ATAC-seq, H3K27ac, H3K4me1) are required for functional enhancer interpretation.
- **`color_palette`**: RColorBrewer palette for loop-type colour assignments in refinement plots (default: `"Paired"`). Controls the rose, donut, and dumbbell chart colours; karyotype heatmaps use fixed red/purple palettes.
- **`hub_percentile`**: Node-degree quantile threshold for connectivity classification (default: `0.95`). The underlying structural hub classification is inherited from the annotation stage and is independent of expression status.
- **`karyo_bin_size`**: Genomic bin width for karyotype heatmaps (default: `1e5`).
- **`species`**: Genome assembly (`"hg38"`, `"hg19"`, `"mm10"`, `"mm9"`). Required for karyotype heatmap chromosome ideograms.

The `eP` and `eG` labels capture expression-aware enhancer-like regulatory states in 3D chromatin space. When available, orthogonal chromatin evidence such as ATAC-seq accessibility, H3K27ac enrichment, or H3K27me3 depletion can further strengthen biological interpretation of these reclassified anchors. `looplook` provides `validate_epeG_by_chromatin()` for systematic testing of eP/eG anchors against user-supplied chromatin mark BED files (see the Orthogonal Validation section below), or the `chromatin_beds` parameter in `refine_loop_anchors_by_expression()` to run validation automatically and add results to the Excel workbook. `refine_loop_anchors_by_chromatin()` provides a complementary chromatin-guided reclassification that updates anchor types and loop topologies; if you plan to use both, skip `chromatin_beds` in the expression refinement step to avoid redundant file reads.

### Refinement Output Structure

The refined `loop_annotation` contains two target gene columns with distinct semantics:

- **`Active_Target_Genes`**: Expression-filtered active-only targets. Contains only genes that pass the expression threshold; no fallback genes are included. This column is used for `Has_Active_Target` and `Retained_In_Functional_Network` determination.
- **`Putative_Target_Genes`**: Display column that may include linear nearest-gene fallback when `Active_Target_Genes` is empty. This preserves backward compatibility with downstream modules.

Each loop is annotated with a **`Refinement_Action`** status:

| Value | Meaning |
|-------|---------|
| `retained_active_target` | Loop has at least one active target gene |
| `reclassified_silent_anchor` | Anchor was reclassified (P→eP or G→eG); no active targets remain |
| `expression_filtered_no_active_target` | Target genes were filtered by expression; no reclassification occurred |
| `structural_only_no_active_target` | Loop had no target genes even before refinement |

The `Has_Active_Target` and `Retained_In_Functional_Network` columns are always identical; both indicate whether the loop belongs to the high-confidence functional subset. To obtain the functional subset, filter on `Retained_In_Functional_Network == TRUE`. When writing to Excel, a dedicated **Functional Loop Annotation** sheet is always included (even if empty) for schema stability.

## Example A: Standard Refinement (Expression Filtering)

In this scenario, a strict expression threshold identifies loops with transcriptionally active target genes. All structural loops are retained; the `Retained_In_Functional_Network` flag indicates which loops belong to the functional subset.

```{r refinement_basic, message=FALSE, warning=FALSE, eval=exists("res_integrated")}
res_basic <- refine_loop_anchors_by_expression(
  annotation_res = res_integrated,
  expr_matrix_file = expr_path,
  sample_columns = c("con1", "con2"),
  threshold = 1.0,
  unit_type = "TPM",
  reclassify_by_expression = FALSE,
  out_dir = out_dir,
  project_name = "Example_Basic_Filter"
)
```

## Example B: Expression-Aware Reclassification (Recommended)

This advanced mode is strongly recommended, as it preserves the valuable 3D structural framework while refining functional annotations. A silent promoter anchor is reclassified as an enhancer (`eP`), acknowledging that while its reference gene is transcriptionally inactive, this region itself may function as an active regulatory element for a distant gene.

```{r refinement_reclassified, message=FALSE, warning=FALSE, eval=exists("res_integrated")}
refined_res <- refine_loop_anchors_by_expression(
  annotation_res = res_integrated,
  expr_matrix_file = expr_path,
  sample_columns = c("con1", "con2"),
  threshold = 1.0,
  unit_type = "TPM",
  reclassify_by_expression = TRUE,
  out_dir = out_dir,
  project_name = "Example_Reclassified_Filter"
)
```

## Example C: Refinement + Orthogonal Chromatin Validation

Supply `chromatin_beds` to automatically validate eP/eG anchors against
chromatin mark data. The result is added to the Excel workbook and returned
as `refined_res$chromatin_validation`.

```{r refinement_chromatin, eval=exists("res_integrated") && exists("out_dir"), message=FALSE, warning=FALSE}
h3k27ac_path <- system.file("extdata", "example_k27ac_peaks.bed", package = "looplook")
h3k4me1_path <- system.file("extdata", "example_h3k4me1_peaks.bed", package = "looplook")
h3k4me3_path <- system.file("extdata", "example_h3k4me3_peaks.bed", package = "looplook")

refined_val <- refine_loop_anchors_by_expression(
  annotation_res = res_integrated,
  expr_matrix_file = expr_path,
  sample_columns = c("con1", "con2"),
  threshold = 1.0,
  reclassify_by_expression = TRUE,
  chromatin_beds = list(
    H3K27ac  = h3k27ac_path,
    H3K4me1  = h3k4me1_path,
    H3K4me3  = h3k4me3_path
  ),
  out_dir = out_dir,
  project_name = "Example_Validated"
)
# View confidence distribution
table(refined_val$chromatin_validation$confidence)
```

### Deep Dive: Refinement Visualizations

This module automatically generates a specialized suite of visual diagnostics to quantify transcriptome-guided refinement efficacy and characterize the functional network:

#### 1. Global Refinement Profiling
*(Always Generated)*

- **Refinement Effect Dumbbell** (`*_Comparison_Dumbbell.pdf`): Quantifies the impact of expression-aware reclassification by visualizing the numerical contrast between original structural loops and refined loops across all topological classes, highlighting reclassification-driven changes in loop type composition.

- **Refined Loop Proportion Rose** (`*_Rose.pdf`): Coxcomb chart illustrating the topological distribution (by count) of the interactome after expression-guided reclassification.
- **Refined Active Genes Karyotype** (`*_Refined_Karyo_Active.pdf`): Genome-wide ideograms mapping the chromosomal density of transcriptionally active loop-associated genes that passed the expression threshold. 

#### 2. Genomic Feature-to-3D Target Annotation Profiling
*(Generated only if `target_bed` was provided in Module 2)*

- **Multi-Omics Sankey Diagram** (`Target_Sankey` htmlwidget in-memory; `target_sankey.html` alongside the report when rendered): Core integration diagnostic that maps the fate of inputted genomic features across three logical tiers: Genomic Region (L1) → 3D Loop Connection (L2) → Target Expression Status (L3). This intuitively reveals the proportion of GWAS/ChIP-seq peaks structurally connected to functionally active genes. When generated through `looplook_report()`, keep `target_sankey.html` in the same output directory as the main HTML report.

- **Refined Target Loop Donut** (`*_Target_Loop_Donut.pdf`): Visualizes the topological distribution of refined structural loops whose anchors overlap user-defined genomic features. This shows the full refined network; for the functional subset, filter on `Retained_In_Functional_Network`.
- **Refined Target Genes Karyotype** (`*_Refined_Karyo_TargetGenes.pdf`): Chromosomal density map of strictly refined putative target genes, facilitating visualization of transcriptionally verified multi-omics loci.

### Orthogonal Validation of eP/eG Anchors (Chromatin Mark Evidence)

After expression-aware refinement produces eP/eG labels, `validate_epeG_by_chromatin()` tests these anchors against user-supplied chromatin mark BED files (ATAC-seq, ChIP-seq) to distinguish genuine regulatory elements from expression artefacts.

**Important:** eP/eG labels are expression-filtered silent states, not functional enhancer calls. This function provides the orthogonal evidence needed to elevate or discard those candidates.

#### Workflow

```{r validate-epeG-example, eval=exists("reclassified"), message=FALSE, warning=FALSE}
h3k27ac_path <- system.file("extdata", "example_k27ac_peaks.bed", package = "looplook")
h3k4me1_path <- system.file("extdata", "example_h3k4me1_peaks.bed", package = "looplook")
h3k4me3_path <- system.file("extdata", "example_h3k4me3_peaks.bed", package = "looplook")

# Validate the eP/eG anchors from the refined result above
val <- validate_epeG_by_chromatin(
  annotation_res = reclassified,
  chromatin_beds = list(
    H3K4me1  = h3k4me1_path,
    H3K27ac  = h3k27ac_path,
    H3K4me3  = h3k4me3_path
  ),
  anchor_gap = 200,          # proximity tolerance (bp)
  anchor_min_overlap = 100   # minimum overlap (bp)
)

# View confidence distribution
table(val$confidence)
```

#### Confidence Criteria

Each eP/eG anchor is tested for overlap with up to 5 marks and assigned a confidence level (highest to lowest):

| Confidence | Condition |
|------------|-----------|
| `gold_standard` | All 5 marks provided; H3K4me1⁺, H3K27ac⁺, ATAC⁺, H3K27me3⁻, H3K4me3⁻ |
| `high_confidence` | H3K4me1⁺ and (H3K27ac⁺ or ATAC⁺); H3K4me3⁻ if tested (H3K4me3⁺ downgrades to `supported`) |
| `supported` | At least one of H3K4me1, H3K27ac, or ATAC is positive. H3K4me3⁺ anchors receive a `promoter_like` tag regardless of H3K27me3 status |
| `weak` | No positive marks, but H3K27me3 or H3K4me3 is tested and absent |
| `uncertain` | No chromatin data provided or no overlaps detected |

Marks not provided are recorded as `NA` and do not influence scoring — missing data is never treated as negative evidence. Anchors with H3K4me3⁺ (with or without H3K27me3⁺) are flagged as `promoter_like`; `refine_loop_anchors_by_chromatin()` reverts these to P (active or poised promoter) or dual-function based on the full mark combination.

#### Output Columns

| Column | Description |
|--------|-------------|
| `anchor_id`, `chr`, `start`, `end` | Anchor identifier and coordinates |
| `anchor_type`, `anchor_gene` | Original type and associated gene(s) |
| `H3K4me1`..`H3K4me3` | `NA` = not tested, `TRUE` = overlap, `FALSE` = tested but absent |
| `confidence` | Factor: `gold_standard` > `high_confidence` > `supported` > `weak` > `uncertain` |
| `evidence` | Human-readable string (e.g. `"H3K4me1+; H3K27ac+; H3K27me3-"`) |

### Chromatin-Guided Anchor Reclassification

While `validate_epeG_by_chromatin()` provides a confidence score for each eP/eG anchor independently, `refine_loop_anchors_by_chromatin()` applies chromatin-mark evidence to systematically **reclassify** anchors and update loop topologies. This function can be called on raw annotation output (`annotate_peaks_and_loops`) to test P/G/E anchors, or on expression-refined output to test eP/eG anchors.

Reclassification rules (minimum input: H3K4me1 + H3K4me3):

- **P + H3K4me1⁺ H3K4me3⁺ → `dual`**: Promoter-enhancer dual-function element
- **P + H3K4me1⁺ H3K4me3⁻ + (H3K27ac⁺ or ATAC⁺) → `E`**: Conservative reclassification requiring active-mark confirmation
- **E + H3K4me3⁺ H3K4me1⁻ → `P`**: Unannotated promoter or ncRNA gene
- **E + H3K4me1⁺ H3K4me3⁺ → `dual`**: Dual-function locus in distal region
- **G + H3K4me1⁺ H3K4me3⁺ → `dual`** / **G + H3K4me3⁺ H3K4me1⁻ → `P`**: Gene-body enhancer or internal promoter
- **eP/eG + promoter_like → `P`/`G`**: Revert to active promoter/gene body
- **eP/eG + dual_like → `dual`**: Confirm dual-function
- Anchor types not matching any rule remain unchanged.

The chromatin state of each anchor is also inferred (highest-priority match first): `conflicting_marks` > `dual_like` > `active_enhancer_like` > `primed_enhancer_like` > `weak_enhancer_like` > `promoter_like` > `repressed` > `uncertain` > `no_data`.

\strong{Important — H3K4me3 dependency:} Rules that test for H3K4me3 *absence* (e.g., P→E, G→E) require H3K4me3 to be provided in `chromatin_beds`. When H3K4me3 is omitted, its value is `NA` for all anchors, and these rules are skipped conservatively — no reclassification without explicit data. Include H3K4me3 (and optionally H3K27me3) to enable the full set of reclassification rules.

```{r chromatin-refine-example, eval=exists("reclassified") && exists("out_dir"), message=FALSE, warning=FALSE}
# Run on expression-refined output (tests eP/eG anchors)
cr <- refine_loop_anchors_by_chromatin(
  annotation_res = reclassified,
  chromatin_beds = list(
    H3K4me1  = h3k4me1_path,
    H3K4me3  = h3k4me3_path,
    H3K27ac  = h3k27ac_path
  ),
  anchor_gap = 200,
  anchor_min_overlap = 100,
  recompute_targets = TRUE,  # Rebuild target links with updated types
  species = "hg38",
  project_name = "Chromatin_Refined",
  out_dir = out_dir
)

# Summary of reclassification
cr$qc_summary

# Updated loop types reflect chromatin evidence
table(cr$loop_annotation$loop_type)

# Chromatin provenance columns in target links
head(cr$target_gene_links[, c("gene", "anchor_type_before_chromatin",
  "anchor_type_after_chromatin", "chromatin_target_action")])
```

The output `chromatin_validation` table records, for each candidate anchor, the overlap with every provided mark (`TRUE`/`FALSE`/`NA`), its confidence level, chromatin state, and the reclassification decision (positional type vs. final type). When `recompute_targets = TRUE`, target gene links carry `anchor_type_before_chromatin`, `anchor_type_after_chromatin`, and `chromatin_target_action` columns for full auditability. When `recompute_targets = FALSE` (default), `target_annotation` and `target_gene_links` are `NULL` — use `profile_target_genes(target_source = "loops")` for chromatin-aware downstream profiling.

#### Chromatin Refinement Visualizations

`refine_loop_anchors_by_chromatin()` automatically generates four diagnostic plots (accessible via `$plots`):

| Plot key | Type | Description |
|----------|------|-------------|
| `Chromatin_Dumbbell` | Dumbbell chart | Anchor-type counts before vs. after reclassification. Grey = original, green = refined. |
| `Chromatin_Sankey` | Sankey flow diagram | Left: anchor types before reclassification. Right: updated types after chromatin-guided reclassification. Green flows = unchanged, amber flows = reclassified. Each flow's width is proportional to the number of anchors. |
| `Chromatin_MarkHeatmap` | Aggregated heatmap | % of anchors in each reclassification group positive for each chromatin mark (white = 0%, deep green = 100%). One row per reclassification type (e.g., P→dual), one column per mark. Left bar: anchor count (N) per group. Cell labels show exact percentages. |
| `Chromatin_Rose` | Rose (coxcomb) chart | Loop-type proportion distribution after chromatin-guided reclassification. |

All plots are returned as **ggplot2** objects (except the heatmap, which is a **ComplexHeatmap** grob) and accept standard `+` layers for customisation. To access a specific plot:

```{r chromatin-plots, eval=exists("cr")}
cr$plots$Chromatin_Dumbbell   # before/after anchor type comparison
cr$plots$Chromatin_Sankey      # reclassification flow diagram
cr$plots$Chromatin_MarkHeatmap # per-anchor mark landscape
cr$plots$Chromatin_Rose        # loop type proportions
```

**Interpreting the MarkHeatmap:** Rows represent reclassification outcomes (e.g., "P → dual" means a promoter was reclassified as a dual-function element). Columns represent chromatin marks. High percentages (dark green) indicate marks that are frequently present in that reclassification group — these are the marks driving the reclassification decision. For example, if "P → dual" shows 95% for H3K4me1 and 88% for H3K4me3, it confirms that dual-mark presence is the primary driver of the P→dual reclassification.

***

## Module 4: Automated Functional Profiling

Following identification of high-confidence putative target genes, `profile_target_genes` provides a fully automated, end-to-end multi-omics analysis pipeline. It integrates 3D genomic interactions with transcriptomic data to systematically characterize the functional landscape and regulatory mechanisms underlying the identified target genes.

\strong{Important:} The GO enrichment, motif scanning, and PPI network modules are designed as \emph{exploratory} (hypothesis-generating) analyses. Statistical tests (Wilcoxon/t-test p-values, GSEA) are reported per-module without global multiple-testing correction across loop types or analysis branches. GSEA uses random down-sampling of target genes controlled by \code{gsea_nSample}, introducing stochasticity. Motif enrichment uses GC-matched background sampling with Fisher's exact test. All exploratory results should be validated with independent experimental approaches before drawing definitive biological conclusions.



### Parameter Strategy: A Highly Modular Pipeline

This function supports highly flexible experimental design through a set of routing parameters that define the gene set to be analyzed, as well as toggle parameters that enable or disable specific downstream biological modules.

#### 1. Core Data Inputs

- **`annotation_res`**: The foundational annotation result, inherited directly from Module 2 or the refined output from Module 3.
- **`diff_file` & `lfc_col`**: Differential expression summary (e.g., from DESeq2) required for fold change profiling and gene set enrichment analysis (GSEA).
- **`expr_matrix_file` & `metadata_file`**: Normalized expression matrix and sample metadata, which are essential for generating heatmaps and Raincloud plots.

#### 2. Target Selection & Mapping Strategies

This section contains the most critical routing parameters that define the biological scope and stringency of downstream analyses:

- **`target_source`** (The Biological Scope):
    - `"targets"` *(Variant/Peak-Centric)*: Focuses exclusively on the putative genes regulated by user-defined genomic features (e.g., GWAS SNPs, ATAC-seq peaks).
    - `"loops"` *(Global Network-Centric)*: Analyzes the full 3D interactome, including all genes associated with spatial loops, without considering user-defined genomic features.
    - `c("loops", "targets")`: Analyzes both scopes simultaneously, which is strongly recommended for comparative functional profiling.
- **`target_mapping_mode`**:
    - `"all"` *(Default)*: Retains broad 3D regulatory targets, including distal enhancers looping to gene bodies.
    - `"promoter"`: Applies stringent filtering, requiring loops to anchor explicitly at canonical promoter regions.
- **`include_Filled`** (The Stringency Toggle): Logical parameter controlling the purity of spatial regulation:
    - `TRUE` *(Hybrid Mode)*: Utilizes the comprehensively merged annotation (`_Filled` columns from Module 2). It prioritizes 3D loop-derived targets while rescuing unlooped genomic features (peaks) by assigning them the nearest genes, ensuring no data loss and a complete functional overview.
    - `FALSE` *(Pure Spatial Mode)*: Highly stringent. It only keeps 3D loop-derived targets with clear annotation.

- **`use_nearest_gene`**: Logical parameter. If `TRUE`, the pipeline bypasses 3D loop-based assignment and uses only the nearest neighboring gene, serving as a conventional baseline reference for Control comparisons.

#### 3. Downstream Functional Analyses

- **`loop_types`**: Character vector of loop types to include in the analysis (default: `c("E-P", "P-P")`). Used only when `target_source = "loops"`.
- **`stat_test`**: Statistical test for comparing target vs. background LFC distributions (`"wilcox.test"` or `"t.test"`; default: `"wilcox.test"`).
- **`org_db`**: Organism annotation database package name for GO and PPI analyses (default: `"org.Hs.eg.db"`).
- **`gsea_nSample`** (numeric, default `99999`): Maximum number of target genes to sample for GSEA. The high default effectively disables down-sampling. When the gene set is large (hundreds to thousands), reduce to e.g. `20–50` to limit enrichment bias introduced by oversized gene sets. Set to `NULL` to always use the full set.
- **`heatmap_nSample`** (numeric, default `99999`): Maximum number of target genes to display in the expression heatmap (top by variance). Reduce to e.g. `20–50` for readable heatmaps.
- **`cnet_nSample`** (numeric, default `50`): Number of top GO terms to display in the concept network plot.
- **`seed`** (integer or `NULL`, default `NULL`): Random seed for reproducible GSEA down-sampling and motif GC-matched background sampling. Set to a positive integer for fully deterministic results.
- **`run_go`**: Executes Gene Ontology enrichment and generates concept network plots (requires `clusterProfiler`).
- **`run_motif` & `genome_id`**: Scans proximal and distal anchor sequences against JASPAR core motifs (requires `BSgenome` and `motifmatchr`).
- **`run_ppi` & `ppi_score`**: Constructs Protein-Protein Interaction (PPI) networks via the STRING database (minimum combined score, default: `400`).

## Example A: Comprehensive Integrative Profiling (Recommended)

Ideal for complete multi-omics integration (e.g., HiChIP + ATAC-seq). This mode comprehensively maps and profiles target genes of interest.

```{r profiling_scenario_A, message=FALSE, warning=FALSE, eval=exists("res_integrated") && requireNamespace("ggpointdensity", quietly=TRUE) && requireNamespace("viridis", quietly=TRUE) && requireNamespace("ggpubr", quietly=TRUE) && requireNamespace("ggdist", quietly=TRUE)}
diff_path <- system.file("extdata", "example_deg.txt", package = "looplook")
meta_path <- system.file("extdata", "example_coldata.txt", package = "looplook")

res_A <- profile_target_genes(
  annotation_res = res_integrated,
  diff_file = diff_path,
  lfc_col = "log2FoldChange",
  expr_matrix_file = expr_path,
  metadata_file = meta_path,
  target_source = c("loops", "targets"),
  target_mapping_mode = "all",
  include_Filled = TRUE,
  use_nearest_gene = FALSE,
  project_name = "Scenario_A_Integrative",
  run_motif = FALSE,
  run_go = FALSE,
  run_ppi = FALSE
)
```

## Example B: Peak-Driven Strict Promoter Profiling (High Stringency)

This mode applies a rigorous dual-filtering strategy to yield a high-confidence target gene set.

First, `target_source = "targets"` restricts profiling to genes regulated specifically by user-defined genomic features (e.g., GWAS SNPs or TF binding sites), excluding background spatial loops. Second, `target_mapping_mode = "promoter"` enforces topological stringency: loops linking peaks to target genes must directly contact canonical or reference promoters (i.e., strict `E-P` or `P-P` loops), excluding more permissive connections such as enhancer-gene body (`E-G`) interactions.

```{r profiling_scenario_B, message=FALSE, warning=FALSE, eval=exists("res_integrated") && requireNamespace("ggpointdensity", quietly=TRUE) && requireNamespace("viridis", quietly=TRUE) && requireNamespace("ggpubr", quietly=TRUE) && requireNamespace("ggdist", quietly=TRUE)}
res_B <- profile_target_genes(
  annotation_res = res_integrated,
  diff_file = diff_path,
  lfc_col = "log2FoldChange",
  expr_matrix_file = expr_path,
  metadata_file = meta_path,
  target_source = "targets", # 1. Focus exclusively on inputted genomic features
  target_mapping_mode = "promoter", # 2. Require strict loop-to-promoter collision
  include_Filled = TRUE,
  use_nearest_gene = FALSE,
  project_name = "Scenario_B_StrictPromoter",
  run_motif = FALSE, # Set TRUE in real analysis to scan JASPAR motifs
  run_go = FALSE, # Set TRUE in real analysis for pathway enrichment
  run_ppi = FALSE
)
```

## Example C: 1D Linear Annotation Baseline (The Control)

By setting `use_nearest_gene = TRUE`, this mode intentionally bypasses the 3D spatial topology and assigns features exclusively to their nearest neighboring gene. It serves as a conventional “baseline control.” Comparing results against Example A or B enables rigorous demonstration of the novel functional insights gained from 3D chromatin looping analyses relative to conventional method.

```{r profiling_scenario_C, message=FALSE, warning=FALSE, eval=exists("res_integrated") && requireNamespace("ggpointdensity", quietly=TRUE) && requireNamespace("viridis", quietly=TRUE) && requireNamespace("ggpubr", quietly=TRUE) && requireNamespace("ggdist", quietly=TRUE)}
res_C <- profile_target_genes(
  annotation_res = res_integrated,
  diff_file = diff_path,
  lfc_col = "log2FoldChange",
  expr_matrix_file = expr_path,
  metadata_file = meta_path,
  target_source = "targets", # Focus on inputted peaks
  target_mapping_mode = "all",
  include_Filled = FALSE,
  use_nearest_gene = TRUE, # Nearest neighboring gene (The Control)
  project_name = "Scenario_C_LinearControl",
  run_motif = FALSE, # Set TRUE in real analysis to scan JASPAR motifs
  run_go = FALSE, # Set TRUE in real analysis for pathway enrichment
  run_ppi = FALSE
)
```

### Deep Dive: Functional Visualizations

Results are returned as a named list, one element per `target_source` (e.g., `"loops"`, `"targets"`). Each element contains `plots` (ggplot objects), `target_gene_sets` (character vectors), and `go_results` (data frames).

```{r profiling-access, eval=exists("res_profile")}
# Accessing results from a profiling run
names(res_profile) # one element per target_source ("loops", "targets")
names(res_profile$loops) # "go_results" "target_gene_sets" "plots"

# --- Target genes ---
res_profile$loops$target_gene_sets # named list of character vectors
names(res_profile$loops$target_gene_sets) # gene set keys (e.g. "All", "Up", "Down")

# --- GO enrichment table ---
go_df <- res_profile$loops$go_results # data frame
head(go_df[
  order(go_df$pvalue), # sort by significance
  c("Description", "ONTOLOGY", "pvalue", "Count", "geneID")
])

# --- All plot keys ---
names(res_profile$loops$plots)
```

#### Available Plot Keys

| Key | Description |
|---|---|
| `LFC_Violin` | Differential expression violin + boxplot |
| `Heatmap` | Expression heatmap with sample annotations |
| `Scatter` | Connectivity vs. expression scatter |
| `Raincloud_LFC` | Raincloud plot of log2 fold changes |
| `Raincloud_Expr` | Raincloud plot of expression levels |
| `GSEA` | Gene Set Enrichment Analysis plot |
| `GO_Network` | GO enrichment cnetplot |
| `GO_Lollipop` | Summary GO lollipop facet |
| `PPI` | STRING protein-protein interaction network |
| `Proximal_Motif_Bar` | Proximal anchor motif enrichment barplot |
| `Proximal_Motif_Logos` | Proximal anchor motif sequence logos |
| `Proximal_Motif_Rank` | Proximal anchor motif rank scatter |
| `Distal_Motif_Bar` | Distal anchor motif enrichment barplot |
| `Distal_Motif_Logos` | Distal anchor motif sequence logos |
| `Distal_Motif_Rank` | Distal anchor motif rank scatter |

All ggplot-based plots accept standard **ggplot2** `+` layers for recolouring or theme adjustment (see Module 2 for examples).

***

## Module 5: IGV-Style Track Visualization

To visualize the local spatial interactome with high resolution, this module generates an IGV-style multi-tiered genomic browser view via the `plot_peaks_interactions` function. This visualization integrates multi-omics datasets into three distinct, non-overlapping genomic tracks, enabling intuitive assessment of 3D structural context:



- **Loop Track (Top)**: Depicts 3D chromatin interactions as Bezier arcs connecting paired spatial anchors. When `score_to_alpha = TRUE` is enabled, the opacity (alpha channel) of each arc is scaled proportionally to the quantitative interaction score.
- **Target Track (Middle)**: Renders user-provided genomic features (e.g., ChIP-seq peaks or GWAS SNPs) from the input BED file, facilitating direct visualization of their spatial overlap with the loop anchors.
- **Gene Track (Bottom)**: Retrieves the longest canonical transcript for each gene using standard Bioconductor annotation databases, with exons and introns mapped according to strand directionality. Overlapping gene models are vertically separated to avoid label collisions and improve readability.

### Parameter Strategy & Core Inputs

The function provides fine-grained control over genomic locus selection, data filtering, and visual aesthetics through the following arguments:

#### 1. Data Inputs & Genomic Coordinates

- **`bedpe_file`**: Character. Specifies the path to the BEDPE-format spatial interaction file. The function utilizes the first six columns for paired anchor coordinates and the seventh column (if present) as the quantitative interaction score.
- **`target_bed`**: Character (Optional). Specifies the path to a user-supplied genomic feature file in standard BED format, rendered in the middle track to assess overlap with 3D anchors.
- **`chr`, `from`, `to`**: Specifies the exact genomic window for visualization. If `chr` is omitted (`NULL`), the function defaults to the chromosome with the highest loop count in the BEDPE file.
- **`min_score`**: Numeric (Optional). Applies a threshold to filter the displayed loops based on the interaction score column.

#### 2. Annotation Dependencies

- **`species`**: Character. Specifies the reference genome assembly (supported: `"hg38"`, `"hg19"`, `"mm10"`, or `"mm9"`). This parameter triggers automatic loading of the corresponding Bioconductor `TxDb` and `OrgDb` annotation packages for gene track rendering.

#### 3. Aesthetic Mapping & Track Configuration

- **`score_to_alpha`**: Logical flag. When `TRUE`, the quantitative interaction score is mapped to the transparency of Bezier arcs, enabling visual differentiation of interaction strength.
- **`max_levels`**: Integer. Defines the maximum number of visible stacking bands for loop arcs. Overlapping loops are separated vertically up to this cap; denser regions are compressed into the available height while preserving relative layering.
- **`base_anchor_height`**: Numeric. Sets the vertical thickness of the rectangular anchors plotted at the base of the loop arcs.
- **Color Controls** (`loop_color`, `anchor_color`, `overlap_color`, `exon_color`, `intron_color`): Character strings specifying custom colors for structural elements across all three tracks.
- **`save_file`**: Character (Optional). Specifies the output file path and format (e.g., `".pdf"`) for exporting the final ggplot object.

```{r visual_track, eval=exists("f1") && exists("atac_path")}
if (requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  # If 'from' and 'to' are omitted, it automatically detects the densest viewport.
  track_plot <- plot_peaks_interactions(
    bedpe_file = f1,
    target_bed = atac_path,
    chr = "chr1",
    from = 11884299,
    to = 12106581,
    species = "hg38",
    save_file = file.path(out_dir, "Locus_Track.pdf")
  )
}
```

***

# One-Click Parameterised Report

For rapid sharing with collaborators, a publication-ready HTML report can be generated with a single function call. The template automatically detects which pipeline stages are configured and hides unavailable sections:

```{r report-example, eval=FALSE}
library(looplook)

# ── Required ──────────────────────────────────────────────────
#   bedpe_file         Chromatin loops in BEDPE format
# ── Required for annotation ───────────────────────────────────
#   target_bed         Genomic features (ATAC-seq peaks, etc.)
# ── Optional annotation tuning ─────────────────────────────────
#   anchor_gap = 0L          Peak-anchor max gap (0 = must touch)
#   anchor_min_overlap = 1L  Min overlap in bp (1 = any touch)
#   anchor_min_frac = 0      Min overlap as fraction of peak width
#   neighbor_hop = 0         k-hop ego-network expansion
# ── Required for refinement + profiling ───────────────────────
#   expr_matrix_file   Normalised expression matrix (TPM/FPKM)
#   diff_file          Differential expression table (CSV/TSV)
#   metadata_file      Sample metadata with Group column
# ── Optional (defaults shown) ─────────────────────────────────
#   species = "hg38"   project_name = "looplook Analysis"
#   threshold = 1.0    unit_type = "TPM"
#   run_go = TRUE      run_ppi = FALSE     run_motif = FALSE
#   ... (see ?looplook_report for full list)
# ── Reproducibility ────────────────────────────────────────
#   The profiling stage uses R's random number generator for GSEA gene-set
#   down-sampling and motif background sampling. For fully reproducible
#   reports, call set.seed() before looplook_report():
#     set.seed(42)
#     looplook_report(...)

# Minimal: annotation only
looplook_report(
  bedpe_file   = system.file("extdata", "example_loops_1.bedpe", package = "looplook"),
  target_bed   = system.file("extdata", "example_peaks.bed", package = "looplook"),
  project_name = "Quick Annotation"
)

# Full pipeline: annotation + refinement + profiling
looplook_report(
  bedpe_file       = "my_loops.bedpe",
  target_bed       = "my_peaks.bed",
  expr_matrix_file = "my_tpm.txt",
  diff_file        = "my_deg.csv",
  metadata_file    = "my_coldata.txt",
  project_name     = "Integrative HiChIP Analysis",
  run_go           = TRUE
)

# Reuse precomputed results: skip annotation and jump to refinement + profiling.
# precomputed_res accepts a file path (.RData) or an in-memory list object.
# bedpe_file is optional when precomputed_res is set.

# Option A — save and reload from .RData:
res <- annotate_peaks_and_loops(
  bedpe_file = "my_loops.bedpe",
  target_bed = "my_peaks.bed",
  project_name = "MyProject"
)
save(res, file = "annotation_results.RData")

looplook_report(
  precomputed_res  = "annotation_results.RData",
  expr_matrix_file = "my_tpm.txt",
  diff_file        = "my_deg.csv",
  metadata_file    = "my_coldata.txt",
  run_go           = TRUE
)

# Option B — pass the object directly:
looplook_report(
  precomputed_res  = res,
  expr_matrix_file = "my_tpm.txt",
  diff_file        = "my_deg.csv",
  metadata_file    = "my_coldata.txt",
  run_go           = TRUE
)
```

The template is also accessible from the RStudio menu: **File → New File → R Markdown → From Template → looplook Report**.

***

# Troubleshooting

## Installation

### TxDb or OrgDb package not found

```
Error: TxDb 'TxDb.Hsapiens.UCSC.hg38.knownGene' not installed
```

**Cause:** The required annotation database package is not installed. These are Bioconductor packages, not CRAN.

**Solution:**
```r
BiocManager::install("TxDb.Hsapiens.UCSC.hg38.knownGene")
BiocManager::install("org.Hs.eg.db")
```

For mouse data:
```r
BiocManager::install("TxDb.Mmusculus.UCSC.mm10.knownGene")
BiocManager::install("org.Mm.eg.db")
```

## Input Data

### BEDPE file must have at least 6 columns

```
Error: BEDPE file must have at least 6 columns.
```

**Cause:** The BEDPE file has fewer than 6 columns. BEDPE format requires: chr1, start1, end1, chr2, start2, end2.

**Solution:** Check that the file is tab-delimited and contains at least 6 columns. Additional columns (name, score, strand) are optional and retained in the output.

### BEDPE file contains rows with start >= end

```
Error: BEDPE file contains rows with start >= end (zero-width or invalid).
```

**Cause:** Some rows have invalid coordinates where the start position is greater than or equal to the end position.

**Solution:** Filter out invalid rows before importing:
```bash
awk '$2 < $3 && $5 < $6' input.bedpe > clean.bedpe
```

### Expression matrix contains non-numeric values

```
Error: Expression matrix contains non-numeric values in sample columns
```

**Cause:** The expression matrix contains text values (e.g., gene names) in sample columns.

**Solution:** Ensure the expression matrix has gene identifiers in the first column and numeric values (TPM, FPKM, counts) in all other columns.

## Gene Name Matching

### Only X% of annotation gene symbols match

```
Warning: Only 12.5% of annotation gene symbols match the expression matrix row names.
```

**Cause:** Gene identifiers in the expression matrix do not match the OrgDb SYMBOL convention. This commonly occurs when:
- Expression matrix uses Ensembl IDs (ENSG00000141510) but OrgDb returns gene symbols (TP53)
- Case mismatch: expression matrix has "tp53" but OrgDb returns "TP53"

**Solution:**
```r
# Check current gene names
head(rownames(expr_matrix))

# Convert to uppercase if needed
rownames(expr_matrix) <- toupper(rownames(expr_matrix))

# Or map Ensembl IDs to symbols
library(org.Hs.eg.db)
symbols <- mapIds(org.Hs.eg.db, keys = rownames(expr_matrix),
                  column = "SYMBOL", keytype = "ENSEMBL")
rownames(expr_matrix) <- symbols
```

**Note:** looplook performs case-insensitive matching internally (using `toupper()` on both sides), but it is best practice to use consistent naming.

## Memory and Performance

### Clustering large datasets

```
Warning: Clustering 50000 loops with gap = 1000 bp. Large datasets may require significant memory.
```

**Cause:** The connected-component clustering algorithm has O(N²) worst-case complexity per chromosome pair.

**Solution:**
```r
# Pre-filter low-confidence loops
consolidate_chromatin_loops(
  files = c("rep1.bedpe", "rep2.bedpe"),
  mode = "consensus",
  min_raw_score = 2,    # Remove singleton noise
  gap = 500,            # Reduce gap for faster clustering
  chaining_policy = "drop"  # Remove chained clusters
)
```

### Chaining warning

```
Warning: 5 cluster(s) have max_span > chaining threshold (5000 bp).
```

**Cause:** Connected-component clustering may transitively merge distant loops through intermediate anchors (A-B + B-C merges A and C).

**Solution:**
```r
# Option 1: Drop chained clusters
consolidate_chromatin_loops(..., chaining_policy = "drop")

# Option 2: Reduce gap to prevent chaining
consolidate_chromatin_loops(..., gap = 500)

# Option 3: Inspect and manually filter
result <- consolidate_chromatin_loops(..., chaining_policy = "none")
# Filter by span
anchors <- anchors(result)
span <- end(anchors$first) - start(anchors$first) + 1
result <- result[span < 5000]
```

## Empty Results

### All loops filtered by min_raw_score

```
Message: All loops filtered out by min_raw_score = 10. Returning empty result.
```

**Cause:** The pre-filtering threshold is too high for the data.

**Solution:** Lower the threshold or disable pre-filtering:
```r
# Check score distribution first
gi <- bedpe_to_gi("my_loops.bedpe")
summary(mcols(gi)$score)

# Use appropriate threshold
consolidate_chromatin_loops(..., min_raw_score = 2)
```

### No peaks overlapped loop anchors

```
Message: No peaks overlapped loop anchors. Check that target BED and loop BEDPE use the same genome build.
```

**Cause:** The target BED file and loop BEDPE file use different genome builds (e.g., hg38 vs hg19) or different chromosome naming conventions (chr1 vs 1).

**Solution:**
```r
# Check chromosome names
seqlevelsStyle(gr_peaks)   # Should match
seqlevelsStyle(gr_anchors)  # Should match

# Harmonize if needed
seqlevelsStyle(gr_peaks) <- "UCSC"  # Convert to chr1 format
```

## Chromatin Refinement

### eP/eG labels dominate the results

```
Warning: 75% of P/G anchors were reclassified to eP/eG.
```

**Cause:** The expression threshold is too high, or the expression matrix does not match the gene annotation.

**Solution:**
```r
# Check expression distribution
hist(log2(vals + 1), main = "Expression Distribution")

# Lower threshold
refine_loop_anchors_by_expression(..., threshold = 0.5)

# Or use quantile mode
refine_loop_anchors_by_expression(..., threshold = 0.25, 
                                   threshold_mode = "quantile")
```

**Important:** eP/eG labels indicate transcriptional silence, not functional enhancer activity. Validate with orthogonal chromatin data (ATAC-seq, H3K27ac) before biological interpretation.

***

# Session Info

```{r session_info, eval = TRUE}
sessionInfo()
```
