---
title: "looplook: An Integrative Framework for Chromatin Interaction Analysis"
author: 
  - name: Ying Zhang
    affiliation: Zhejiang University
    email: 12207129@zju.edu.cn
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{looplook: An Integrative Framework for Chromatin Interaction Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Global setup: 
# eval = FALSE ensures the document builds even if example files are missing.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 6,
  eval = FALSE  
)

```

# Introduction

**looplook** is a comprehensive R/Bioconductor toolkit designed to bridge the gap between 3D genomic interactions (HiChIP, ChIA-PET, Hi-C) and transcriptional landscapes. It provides a seamless workflow for:

1. **Consolidating** biological replicates using graph-based clustering.
2. **Annotating** loops with auxiliary genomic features (e.g., ATAC-seq peaks).
3. **Refining** regulatory definitions using RNA-seq data (filtering inactive promoters).
4. **Profiling** downstream functions via automated GO/GSEA and PPI networks.

This vignette provides a step-by-step guide to the `looplook` pipeline.

# Installation

```{r install}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

# Install from GitHub
devtools::install_github("your_github_username/looplook")

# Load package
library(looplook)

```

# Module 1: Data Preprocessing & Consolidation

## Importing Data

The `bedpe_to_gi` function imports standard BEDPE files into Bioconductor `GInteractions` objects, automatically normalizing anchor order and detecting score columns.

```{r read_data}
bedpe_file <- system.file("extdata", "example_loops_1.bedpe", package = "looplook")
gi_obj <- bedpe_to_gi(bedpe_file)

```

## Consolidating Replicates

Use `consolidate_chromatin_loops` to merge biological or technical replicates and define a high-confidence consensus loop set.

### Key Parameters

| Parameter | Description | Recommendation |
| --- | --- | --- |
| `files` | Character vector of BEDPE file paths. | Required (). |
| `mode` | Strategy for merging: <br>

<br>• **`reproducible`**: Graph clustering (keeps clusters in  reps).<br>

<br>• **`intersect`**: Strict overlap filtering.<br>

<br>• **`union`**: Merges all loops. | Use `"reproducible"` for biological replicates. |
| `gap` | Max distance (bp) to merge anchors. | Default `1000` bp. |
| `min_reproducible` | Min reps required to keep a loop. | Default `NULL` (auto-majority rule). |

### Example Usage

```{r consolidate}
f1 <- system.file("extdata", "example_loops_1.bedpe", package = "looplook")
f2 <- system.file("extdata", "example_loops_2.bedpe", package = "looplook")

consensus_loops <- consolidate_chromatin_loops(
  files = c(f1, f2),
  mode = "reproducible",
  gap = 1000,
  min_reproducible = 2,
  out_file = "results/consensus_loops.bedpe"
)

```

# Module 2: Annotation & Integration

## Basic Annotation

Classify loops (e.g., Promoter-Promoter, Enhancer-Promoter) and integrate auxiliary data (e.g., ATAC-seq peaks) to link open chromatin regions to target genes via 3D loops.

### Key Parameters

| Parameter | Description |
| --- | --- |
| `target_bed` | Path to auxiliary BED file (e.g., ATAC-seq, GWAS SNPs). |
| `species` | Genome build: `"hg38"`, `"hg19"`, `"mm10"`, `"mm9"`. |
| `tss_region` | Window around TSS. Default `c(-2000, 2000)`. |

### Outputs

* **Excel**: `_Basic_Results.xlsx` (Loop types, Cluster hubs, Target links).
* **Plots**: Rose Plot, Circular Bar Plot, Karyotype Heatmaps.

```{r annotation}
loops_path <- "results/consensus_loops.bedpe"
atac_path <- system.file("extdata", "example_peaks.bed", package = "looplook")

anno_res <- annotate_peaks_and_loops(
  bedpe_file = loops_path,
  target_bed = atac_path,
  species = "hg38",
  out_dir = "results/annotation",
  project_name = "Study_H3K27ac"
)

```

# Module 3: Expression-Based Refinement

Standard annotation relies on distance. `refine_loop_anchors_by_expression` incorporates RNA-seq to filter false positives: **inactive promoters are reclassified as enhancers**.

### Key Parameters

| Parameter | Description |
| --- | --- |
| `expr_matrix_file` | Path to expression matrix (TPM/FPKM). |
| `sample_columns` | Columns to average (e.g., `c("Rep1", "Rep2")`). |
| `threshold` | Expression cutoff (default `1.0`). |
| `downgrade_nonexpressed` | Logical (`TRUE`). Reclassify silent promoters as enhancers. |

```{r refinement}
expr_file <- "data/tpm_matrix.tsv" 

refined_res <- refine_loop_anchors_by_expression(
  annotation_res = anno_res,
  expr_matrix_file = expr_file,
  sample_columns = c("Rep1", "Rep2"),
  threshold = 1.0,
  out_dir = "results/refined",
  project_name = "Study_Refined"
)

```

# Module 4: Functional Profiling

The `profile_target_genes` pipeline performs multi-omics characterization:

1. **LFC Profiling**: Target genes vs Background (Boxplots).
2. **Enrichment**: GO & GSEA Analysis.
3. **Clustering**: Heatmaps & Correlation.
4. **Networks**: PPI construction via STRING.

```{r profiling}
profile_target_genes(
  annotation_res = refined_res,
  use_filtered = TRUE,
  diff_file = "data/deseq2_results.csv",
  lfc_col = "log2FoldChange",
  tpm_file = expr_file,
  metadata_file = "data/metadata.csv",
  target_source = "cluster",
  loop_types = c("EP"), 
  out_dir = "results/profiling"
)

```

# Module 5: Visualization

## Integrative Track Plot

Generate publication-quality IGV-style plots with arcs (loops) and blocks (genes/peaks).

```{r visual_track}
plot_peaks_interactions(
  bedpe_file = loops_path,
  target_bed = atac_path,
  chr = "chr8",
  from = 127730000,
  to = 127745000,
  species = "hg38",
  save_file = "results/plots/MYC_locus.pdf"
)

```

## Flower Plot

Visualize gene set intersections (Core vs Unique).

```{r visual_stats}
draw_flower_simplified(
  gene_lists = list(Ctrl = c("A","B","C"), Treat = c("B","C","D")),
  project_name = "Comparison",
  filename = "results/plots/flower.pdf",
  group_colors = c("#E41A1C", "#377EB8")
)

```

# Session Info

Here is the output of `sessionInfo()` on the system on which this document was compiled:

```{r session_info, eval = TRUE}
sessionInfo()

```

```

```
